Structure and evolution of Alanine/Serine Decarboxylases and the engineering of theanine production

Version 1

Abstract

Abstract

Ethylamine (EA), the precursor of theanine biosynthesis, is synthesized from alanine decarboxylation by Alanine Decarboxylase (AlaDC) in tea plants. AlaDC evolves from Serine Decarboxylase (SerDC) through neofunctionalization and has lower catalytic activity. However, lacking structure information hinders the understanding of the evolution of substrate specificity and catalytic activity. In this study, we solved the X-ray crystal structures of AlaDC from Camellia sinensis (CsAlaDC) and SerDC from Arabidopsis thaliana (AtSerDC). Tyr341 of AtSerDC or the corresponding Tyr336 of CsAlaDC is essential for their enzymatic activity. Tyr111 of AtSerDC and the corresponding Phe106 of CsAlaDC determine their substrate specificity. Both CsAlaDC and AtSerDC have a distinctive zinc finger that contributes to the activity and has not been identified in any other Group II PLP-dependent amino acid decarboxylases. Based on the structure comparison and evolution analysis AlaDC or SerDC led us to perform mutation screen of CsAlaDC. The results indicated that the mutation of L110F or P114A in the CsAlaDC dimerization interface significantly improved the catalytic activity by 110% and 59%, respectively. Combining double mutant CsAlaDCL110F/P114A with theanine synthetase increased 672% of theanine production in an in vitro system. This study provides the structural basis for the substrate selectivity and catalytic activity of CsAlaDC and AtSerDC and sheds light on a new direction for the efficient industrial synthesis of theanine.

Introduction

Tea is one of the most popular natural non-alcoholic beverages consumed worldwide. About 2 billion cups of tea are consumed worldwide daily (1). The popularity of tea is determined by its favorable flavor and numerous health benefits (2–4). The flavor and health-beneficial effects of tea are conferred by the abundant secondary metabolites, including catechins, caffeine, theanine, volatiles, etc (5). Theanine (γ-glutamylethylamide) is a unique non-protein amino acid and the most abundant free amino acid in tea plants (Camellia sinensis). It accounts for more than 50% of the total free amino acids and approximately 1-2% of the dry weight of the new shoots of tea plants (6). Theanine is the secondary metabolite conferring the umami taste of tea infusion and also balances the astringency and bitterness of tea infusion caused by catechins and caffeine (7). It has also many health-promoting functions, such as reducing stress and anxiety, improving mood and cognition, protecting the nervous system, etc (6, 8–13). Therefore, theanine content is highly correlated with green tea quality (14).

Theanine is synthesized from EA and glutamate (Glu) by theanine synthetase (15, 16). Importantly, the large amount of theanine biosynthesis is determined by the high availability of EA in tea plants (17). Therefore, the evolution of EA biosynthesis in tea plants largely determined the quality formation of tea. EA is synthesized from alanine decarboxylation by alanine decarboxylase (CsAlaDC) (18) (Figure 1A). Indeed, CsAlaDC expression level and catalytic activity largely determine the theanine accumulation level in tea plants (19). As a novel gene, CsAlaDC had not been reported in other plant species before it was identified in tea plants. Previous studies indicated that AlaDC originated from the serine decarboxylase gene (SerDC) by gene duplication in tea plants (20). However, CsAlaDC and CsSerDC specifically catalyze alanine and serine decarboxylation, respectively, despite they share highly conserved amino acid sequences (20) (Figure 1A, 1C). In addition, CsAlaDC exhibited a much lower enzymatic activity compared with CsSerDC. However, the structural basis and key amino acids underlying the evolution of the substrate specificity and enzymatic activity of CsAlaDC are unknown.

Metabolic pathways and phylogeny analysis of SerDC and AlaDC in plants.

(A) The decarboxylation of serine and alanine in plants. SerDC, serine decarboxylase; AlaDC, alanine decarboxylase. (B) Phylogenetic analysis of the SerDC homologous proteins in Algal, Pteridohyta, Bryophyta, Dicots and Monocots. The phylogenetic tree was performed using the neighbor-joining method in MAGA 7.0 software and bootstrap values based on 1000 replicates were displayed at branch points. These proteins were represented by their GenBank accession number. (C) Multiple alignment of the amino acid sequences of AlaDC and SerDCs. Primary (100%), secondary (80%), and tertiary (60%) conserved percent of similar amino acid residues were shaded in deep blue, light blue and cheer red, respectively. The multiple alignment was constructed by MAGA 7.0 software by ClustalW method, and GeneDOC software was used to visualize multiple sequence alignments. “*” indicated amino acid residues mutated only in CsAlaDC.

SerDC belongs to the Group II pyridoxal-5-phosphate (PLP)-dependent decarboxylase superfamily (21). These Group II amino acid decarboxylases produce many bioactive secondary metabolites and signaling molecules (22–24). In plants, SerDC catalyzes the biosynthesis of ethanolamine from serine, which was first functionally characterized in Arabidopsis thaliana (25) (Figure 1A). Ethanolamine is an important metabolite in the synthesis of phosphatidylethanolamine and phosphatidylcholine, two main phospholipids involved in maintaining eukaryotic membrane structures in eukaryotic cell membranes (26–28). Furthermore, the study has shown that SerDC is essential in the embryogenesis of Arabidopsis (29). SerDC is of vital importance for plant growth and development, but the mechanism of substrate recognition and catalytic activity of SerDC also remains unclear.

In this work, we attempt to understand the mechanism of functional divergence of plant AlaDC and SerDC by analyzing their structural characteristics. Here, we obtained the X-ray crystal structure of CsAlaDC and AtSerDC. According to the crystal structure, we found a distinctive zinc finger structure located at both CsAlaDC and AtSerDC, which has not been identified in any other Group II PLP-dependent amino acid decarboxylases that have been previously characterized. By comparing the substrate binding pockets, we identified Phe106 of CsAlaDC and Tyr111 of AtSerDC as the crucial sites for substrate specificity. By conducting mutation screening based on the protein structures, we identified the amino acids repressing the catalytic activity and discovered that CsAlaDCL110F/P114A exhibited a 2.3-fold increase in catalytic activity compared to that of CsAlaDC and enhanced the engineering of theanine production in vitro.

Results

Enzymatic properties of CsAlaDC, AtSerDC, and CsSerDC

CsAlaDC originates from CsSerDC by gene duplication and neofunctionalization in tea plants, but they catalyze different metabolic processes (Figure 1A). A phylogenetic tree was constructed to further reveal the evolutionary relationship between CsAlaDC and its homologs in Algal, Pteridohyta, Bryophyta, Monocots and Dicots (Figure 1B). Phylogenetic analysis indicated that CsAlaDC is homologous with SerDCs, and its closer relationship was Dicots plants. As expected, CsSerDC was most closed to AtSerDC, which implies that they shared similar functions. However, CsAlaDC is relatively distant from CsSerDC. Next, we performed a multiple sequence alignment of the amino acid sequences of the 6 SerDCs and CsAlaDC (Figure 1C). The results showed that the amino acid sequences of CsAlaDC and SerDCs were highly conserved, but CsAlaDC has amino acid mutations in conserved motifs compared with other SerDCs.

Next, to verify the substrate specificity and enzyme activity, we conducted enzyme activity detection and enzyme kinetics analysis for CsAlaDC, CsSerDC and AtSerDC. The 5′ truncated CsAlaDC and AtSerDC were constructed into the pET22b and pET28a expression vectors to generate recombinant plasmids pET22b-CsAlaDC and pET28a-AtSerDC, correspondingly. Additionally, the full-length protein of CsSerDC was inserted into the pET28a expression vector to generate recombinant plasmid pET28a-CsSerDC. Subsequently, the recombinant proteins were expressed in Escherichia coli and purified via nickel affinity chromatography. The purified proteins were examined through SDS-PAGE analysis (Figure 2A). To investigate the substrate specificity of CsAlaDC, AtSerDC, and CsSerDC, enzyme activity assays were conducted using Ala and Ser as substrates, followed by product identification via UPLC analysis. The results showed that CsAlaDC can effectively participate in the decarboxylation of alanine to generate EA, whereas its catalytic efficacy towards serine is relatively inferior. Conversely, AtSerDC and CsSerDC selectively catalyzed serine decarboxylation to yield ethanolamine but did not exhibit activity towards alanine (Figure 2B). These findings confirmed that CsAlaDC and SerDCs do not have promiscuous decarboxylase activity on Ala and Ser.

Purification and characterization of CsAlaDC, AtSerDC, and CsSerDC.

(A) Identification of CsAlaDC, AtSerDC, and CsSerDC using SDS-PAGE. (B) Detection of enzyme activities of CsAlaDC, AtSerDC, and CsSerDC by UPLC. (C) Reaction rates of substrates with different concentrations catalyzed by CsAlaDC, AtSerDC, and CsSerDC.

The kinetic properties of CsAlaDC, AtSerDC, and CsSerDC were determined through the use of corresponding substrates (Figure 2C), and the results are presented in Table 1. The kinetic parameter Km of CsAlaDC was determined to be 1.215 mmol/L, which is similar to that of AtSerDC and CsSerDC at 1.522 mmol/L and 2.364 mmol/L, respectively. However, AtSerDC has a Vmax of 7.053 μmol/L/s, which is 4-fold that of CsAlaDC’s Vmax of 1.709 μmol/L/s; while CsSerDC has a Vmax of 9.031 μmol/L/s, which is 5.3-fold that of CsAlaDC. These findings indicated that the catalytic efficiency of CsAlaDC is considerably lower than that of both CsSerDC and AtSerDC. Due to the similarity in catalytic activity between CsSerDC and AtSerDC, we chose the more representative AtSerDC for further analysis.

The overall structures of CsAlaDC and AtSerDC

To enhance our comprehension of CsAlaDC and AtSerDC, we conducted structural analyses of these two proteins. Through optimization of crystallization conditions, we successfully determined the crystal structures of CsAlaDC, as well as the CsAlaDC-EA, and AtSerDC at resolutions of 2.50 Å, 2.60 Å and 2.85 Å, respectively (Supplementary table 1).

The overall architecture of CsAlaDC and AtSerDC comprises homodimeric structures with two subunits exhibiting an asymmetrical arrangement (Figure 3A and 3D). The monomer of CsAlaDC and AtSerDC is divided into three distinct structural domains: an N-terminal domain (N-terminal −104 aa in CsAlaDC, N-terminal −109 aa in AtSerDC), a large domain (105-364 aa in CsAlaDC, 110-369 aa in AtSerDC), and a small C-terminal domain (365 aa-C-terminus in CsAlaDC, 370-C-terminus as in AtSerDC), colored light pink, khaki, and sky blue, respectively. Compared to the full-length protein, the N-terminal structures of CsAlaDC and AtSerDC are truncated by 60 and 65 amino acid residues, respectively. The truncated protein N-terminus contains a long α-helix, which is connected to the large domain through a long loop. The large domain contains a seven-stranded mixed β-sheet surrounded by eight α-helices, while the C-terminal domain composes of a three-stranded antiparallel β-sheet and three α-helices. The catalytic site of the enzyme is positioned within a superficial crevice at the junction of two subunits forming a dimeric arrangement. The cofactor-binding process involves amino acid residues derived from both subunits. One monomer accommodates both PLP and substrate, while the other monomer also contributes to cofactor binding and enzymatic function (Figure 3A and 3D).

Crystal structures of CsAlaDC and AtSerDC.

(A) Dimer structure of CsAlaDC. The color display of the N-terminal domain, large domain, and C-terminal domains of chain A is shown in light pink, khaki and sky blue, respectively. Chain B is shown in spring green. The PLP molecule is shown as a sphere model. The zinc finger structure at the C-terminus of CsAlaDC is indicated by the red box. The gray spheres represent zinc ions, while the red dotted line depicts the coordination bonds formed by zinc ions with cysteine and histidine. (B) The 2Fo-Fc electron density maps of K309-PLP-EA (contoured at 1σ level). The PLP is shown in violet, the K309 is shown in spring green, and the EA is shown in lightblue. (C) Active center of the CsAlaDC-EA complex, with hydrogen bonds denoted by black dotted lines. “*” denotes the amino acids on adjacent subunits. (D) Dimer structure of AtSerDC. The color display of the N-terminal domain, large domain, and C-terminal domains of chain A is shown in light pink, khaki and sky blue, respectively. Chain B is shown in cyan. The PLP molecule is shown as a sphere model. The zinc finger structure at the C-terminus of AtSerDC is indicated by the red box. The gray spheres represent zinc ions, while the red dotted line depicts the coordination bonds formed by zinc ions with cysteine and histidine. (E) Active center of the AtSerDC, with hydrogen bonds denoted by black dotted lines. “*” denotes the amino acids on adjacent subunits. (F) The monomer of CsAlaDC and AtSerDC are superimposed. CsAlaDC is depicted in spring green, while AtSerDC is shown in cyan. The conserved amino acid catalytic ring is indicated by the red box (G) Amino acid residues of the active center in CsAlaDC apo and CsAlaDC-EA complex are superimposed. CsAlaDC apo is shown in floral white, while CsAlaDC-EA complex is shown in spring green. (H) The relative activity of wild-type CsAlaDC and its Y336F mutant (left), as well as wild-type AtSerDC and its Y341F mutant (right) is shown. (I) The effects of DTT on the activity of CsAlaDC and AtSerDC are depicted. Three independent experiments were conducted.

Notably, our investigation has revealed the presence of a distinctive zinc finger structure (as depicted in Figure 3A and 3D) located at the C-terminus of CsAlaDC and AtSerDC. This structure is composed of a ring structure spanning 17 amino acid residues, wherein coordination of Zn2+ is facilitated by three Cys residues and one His residue. Importantly, this particular configuration is exclusive to the two proteins under examination and has not been identified in any other Group II PLP-dependent amino acid decarboxylases that have been previously characterized.

The crystal structures of CsAlaDC-EA complex and AtSerDC were obtained and further analyzed. In the former complex, PLP bound to Lys309 via a Schiff base linkage (internal aldimine form), and the pyridine moiety of PLP is positioned between the imidazole ring of His196 and the methyl group of Ala279 in a parallel orientation to the imidazole ring. The N1 of PLP establishes a salt bridge interaction with Asp277, while the O3 forms hydrogen bonds with Thr247 and Lys309. Additionally, the phosphate group of PLP participates in hydrogen bonding interactions with His308, Gly169, Thr170, Lys309, and Ser347*(“*” denotes the amino acids on adjacent subunits.) (Figure 3C). Similarly, in the latter structure, the pyridine ring of PLP was sandwiched between the imidazole ring of His201 and the methyl group of Ala284. The N1 of PLP formed a salt bridge with Asp282, while the O3 formed hydrogen bonds with Thr252. The phosphate group of PLP established hydrogen bonding interactions with Lys314, His313, Gly174, Thr175, and Ser352* (Figure 3E).

Factors affecting CsAlaDC and AtSerDC activity

The Group II PLP-dependent amino acid decarboxylases are dimeric enzymes with each monomer containing an active site located within a shallow cavity (30). A flexible loop of amino acids originating from one monomer extends into the active site of the other monomer and plays a crucial role in catalysis. The catalytic loop in CsAlaDC, consisting of amino acid residues 328-341, is well-structured and exhibits clear electron density. However, the corresponding loop in AtSerDC is disordered (Figure 3F). The aforementioned loop harbors a conserved Tyr residue that is believed to function as a proton donor of the carbanion in the catalytic process (31). In particular, CsAlaDC Tyr336 and AtSerDC Tyr341 correspond to that Tyr residue. Upon superimposing the active site residues of CsAlaDC apo and CsAlaDC-EA complex, it was observed that the hydroxyl group of Tyr336* in the CsAlaDC-EA complex exhibits a substantial 60-degree deviation along the Cα-Cβ bound (Figure 3G). To assess the function of the Tyr, we substituted the corresponding Tyr residues in CsAlaDC and AtSerDC with Phe. The resulting CsAlaDCY336F and AtSerDCY341F mutants were then exposed to L-alanine and L-serine, respectively. We measured the production of EA or ethanolamine in the reaction mixtures. Our findings indicate that these mutants catalyze abortive decarboxylation, as opposed to wild-type CsAlaDC and AtSerDC, as evidenced by the absence of detectable EA and only a small amount of ethanolamine observed in the reaction mixture (Figure 3H). This result suggested this Tyr is required for the catalytic activity of CsAlaDC and AtSerDC.

CsAlaDC and AtSerDC both possess a C-terminal zinc finger motif whose removal led to protein insoluble when expressed in E. coli. Therefore we used dithiothreitol (DTT) to block the zinc finger structure to evaluate the role of this zinc finger structure in the catalytic activity. The results showed that 5 mM L-DTT reduced the relative activity of CsAlaDC and AtSerDC to 22.0% and 35.2%, respectively, compared to the control (Figure 3I). These results suggested that this characteristic zinc finger motif is probably critical for the protein folding of CsAlaDC and AtSerDC.

Identification of key amino acids for the substrate specificity

Through the superposition of the amino acid residues present in the substrate binding pocket of CsAlaDC and AtSerDC, it was observed that except for Phe at position 106 in CsAlaDC (Figure 4A), the residues were identical. In AtSerDC, this position is occupied by Tyr at 111, suggesting that this specific amino acid residue is related to substrate specificity.

Key amino acid residues for substrate recognition.

(A) Superposition of substrate binding pocket amino acid residues in CsAlaDC and AtSerDC. The amino acid residues of CsAlaDC are shown in spring green, the amino acid residues of AtSerDC are shown in cyan, with the substrate specificity-related amino acid residue highlighted in a red ellipse. (B) Active-site-lining residues from Embryophyta AlaDC were identified. The height of the residue label displays the relative amino acid frequency. (C) Histogram showing the distribution of the number of key motifs. (D) Histogram showing the number of key motifs in different plant orders. (E) Relative enzyme activities of wild-type CsAlaDC and mutant protein CsAlaDCF106Y against Ala substrate (columns 1 and 2), and enzyme activities of wild-type AtSerDC and various AtSerDC mutant proteins against Ala substrate (columns 3-9) are presented. The percentage graph shows the relative activity of each protein compared to wild-type CsAlaDC activity (taken as a 100% benchmark). (F) Relative enzyme activities of wild-type AtSerDC and AtSerDC mutant proteins (columns 1-7) against Ser substrates, and enzyme activities of wild-type CsAlaDC and mutant protein CsAlaDCF106Y against Ser substrates(columns 8, 9) were measured. The percentage graph shows the relative activity of each protein compared to the wild-type AtSerDC (taken as a 100% benchmark). Three independent experiments were conducted. (G) The EA contents of AtSerDC and its mutant AtSerDCY111F in N. benthamiana. (H) The EA contents of CsAlaDC and its mutant CsAlaDCF106Y in N. benthamiana. The significance of the difference (P<0.05) was labeled with different letters according to Duncan’s multiple range test.

To obtain more insights into the role of specific amino acid residue in substrate specificity, we utilized amino acid sequences of AtSerDC to identify all potential serine decarboxylases in Embryophyta and constructed phylogenetic trees (Supplementary figure 1). We further identified conserved motifs in these SerDC homologs in Embryophyta (Figure 4B). Importantly, the Phe106 of CsAlaDC or Tyr111 of AtSerDC locates in a conserved motif (Figure 4B). In this conserved motif, the first two amino acids Y and P are 100% conserved, but the third amino acid is variable and can be Y, T, F, V, A, I, or other amino acids (Figure 4B and C); interestingly, the Phe106 of CsAlaDC and Tyr111 of AtSerDC locates at this variable amino acid position. For these first three amino acids, the YPY was identified in 83.7% of the SerDC homologs, while the YPF motif was found in 2.3% of these SerDC homologs (Figure 4C). These analyses provided more insights into the role of Phe106 of CsAlaDC and Tyr111 of AtSerDC in the substrate specificity.

To verify the role of Phe106 of CsAlaDC or Tyr111 of AtSerDC in substrate specificity, we mutated Phe106 of CsAlaDC into Tyr (CsAlaDCF106Y) and Tyr111 of AtSerDC into Phe (AtSerDCY111F) and tested enzymatic activity in vitro. The results showed that CsAlaDCF106Y totally lost the alanine decarboxylase activity, and AtSerDCY111F gained alanine decarboxylase activity (Figure 4E). We also performed other mutations of Tyr111 of AtSerDC, including AtSerDCY111A, AtSerDCY111I, AtSerDCY111L, AtSerDCY111V and AtSerDCY111W. Within these mutations, only AtSerDCY111I exhibited a low level of alanine decarboxylase activity (Figure 4E). These results verified that the Phe106 of CsAlaDC is required for the alanine decarboxylase activity from serine decarboxylase activity.

On the other side, both CsAlaDC and CsAlaDCF106Y exhibit a very low level of serine decarboxylase activity (Figure 4F). Moreover, compared with AtSerDC, the serine decarboxylase activity of AtSerDCY111F was about 30% of AtSerDC; AtSerDCY111I retained lower than 5% of serine decarboxylase activity of AtSerDC; while other mutations, including AtSerDCY111A, AtSerDCY111L, AtSerDCY111V and AtSerDCY111W abolished the serine decarboxylase activity (Figure 4F). These results suggested the Tyr111 of AtSerDC is also important for the substrate specificity of AtSerDC.

To further verify that Phe106 of CsAlaDC and Tyr111 of AtSerDC were key amino acid residues determining its substrate recognition in planta, Nicotiana benthamiana transient expression system was carried out. To this end, A. tumefaciens strain GV3101 (pSoup-p19), carrying recombinant plasmid, was infiltrated into leaves of 5-week-old N. benthamiana plants, and the pCAMBIA1305 empty vector was used as the control (EV). Relative mRNA levels were detected (Supplementary figure 2), indicating that they had been successfully overexpressed in tobacco. Here, we found that a high level of EA was produced in the CsAlaDC-expressing tobacco leaves; while no EA product was detected in tobacco leaves infiltrated with the mutant CsAlaDCF106Y. In addition, we did not detect EA products in the AtSerDC-expressing tobacco leaves, while a high level of EA was detected in tobacco leaves infiltrated with mutant AtSerDCY111F. As anticipated, EA was not detected in tobacco leaves infiltrated with EV (Figure 4G and H). These results further verified the critical role of Phe106 in the substrate specificity of CsAlaDC.

Key amino acids for the evolution of CsAlaDC enzymatic activity

CsAlaDC and AtSerDC have a high sequence similarity of 74.5% and a nearly identical structure with an RMSD (root mean square deviation) of only 0.77Å for monomer structures. However, the two enzymes catalyze different amino acid decarboxylation reactions, and AtSerDC exhibits significantly higher activity than CsAlaDC (Figure 2B and C). To elucidate this observation, we proposed a prediction test: EA, generated via Ala decarboxylation by CsAlaDC, can be toxic and harmful to plants if accumulated excessively. Thus, during the evolution of plant serine decarboxylase into alanine decarboxylase, the enzyme has evolved not just to alter the substrate preference but also to reduce catalytic activity to control EA production within a suitable range. Based on this hypothesis, we suggest that mutating specific amino acids in CsAlaDC to those corresponding amino acids in SerDCs could enhance its activity, and the results could provide insights into the evolution of CsAlaDC enzymatic activity.

Dimerization is essential to AlaDC/SerDC activities because the active site is composed of residues from two monomers. To identify the amino acids repressing the enzymatic activity of CsAlaDC during evolution from SerDC, we analyzed the crystal structures and the amino acids at the dimer interface between CsAlaDC and AtSerDC. This analysis revealed that the amino acid residues at positions 66, 97, 110, 114, 116, 117, 122, 315 and 345 are different in CsAlaDC and SerDCs (Figure 1C). Therefore, we mutated these amino acids of CsAlaDC into those in the corresponding positions of AtSerDC and CsSerDC, and performed enzyme activity assays (Figure 5A). The results demonstrated that CsAlaDCL110F and CsAlaDCP114A exhibited significantly enhanced enzyme activity compared to the wild-type CsAlaDC, with a 2.1-fold and 1.59-fold increase, respectively. Furthermore, the catalytic activity of the CsAlaDCL110F/P114A double mutant exhibits a remarkable 2.3-fold enhancement compared to that of the wild-type protein (Figure 5B). These findings suggested a critical role of Leu110 and Pro114 of CsAlaDC in the evolution of enzymatic activity and provided a basis to improve CsAlaDC activity. It is possible that these amino acid residues could potentially augment the hydrophobic nature of the protein dimer interface.

Mutations enhance CsAlaDC enzyme activity and theanine synthesis in vitro.

(A) Relative enzyme activities of CsAlaDC mutant proteins against Ala substrate. (B) Relative enzyme activities of CsAlaDCL110F, CsAlaDCP114A, and CsAlaDCL110F/P114A against Ala substrate. (C) Histogram showing the relative content of theanine resulting from different combinations of alanine decarboxylase and theanine synthetase. Three independent experiments were conducted.

In Vitro synthesis of theanine

The biosynthetic pathway of theanine in tea plants comprises two sequential enzymatic steps: alanine decarboxylase mediates alanine decarboxylation to yield EA, which is subsequently utilized by theanine synthetase to effectuate the condensation reaction with glutamate to synthesize theanine. By conducting mutation screening, we discovered a CsAlaDC mutant protein (L110F/P114A) that exhibited a 2.3-fold increase in catalytic activity compared to the wild-type protein. Subsequently, we employed an in vitro theanine synthesis system utilizing CsAlaDC and either glutamate synthetase (PtGS [Pseudomonas syringae pv. syringae]) or gamma-glutamate methylamine ligase (GMAS [Methylovorus mays]) which have the activity to synthesize theanine from EA and Glu (32, 33).

The results illustrated the successful synthesis of theanine using CsAlaDC in conjunction with two theanine synthetases, employing Ala and glutamate as substrates (Figure 5C). The theanine content generated by the combination of CsAlaDCL110F and GMAS, as well as CsAlaDCL110F/P114A and GMAS, are 4.57-fold and 6.72-fold higher than the content produced in the combination of wild-type CsAlaDC and GMAS, respectively. Similarly, when combined with the PsGS, comparable outcomes are observed as well. The theanine content resulting from the combination of CsAlaDCL110F and PsGS, as well as CsAlaDCL110F/P114A and PsGS, exhibit enhancements of 1.62-fold and 4.33-fold compared to the wild-type protein combination, respectively. (Figure 5C). Hence, the utilization of CsAlaDCL110F/P114A could effectively enhance the theanine production yield and thus, holds potential for large-scale engineering production of theanine.

Discussion

Comparison of CsAlaDC and AtSerDC with other decarboxylases

The Dali search (34) revealed that AtSerDC and CsAlaDC exhibit structural similarity to several other decarboxylases, including 7CIG (MetDC [Streptomyces sp. 590]), 7ERV (HisDC1 [Photobacterium phosphoreum]), 6KHO (TrpDC [Oryza sativa Japonica Group]), 4E1O (HisDC2 [Homo sapiens]), 6JY1 (AspDC [Methanocaldococcus jannaschii]), and 5GP4 (GluDC [Levilactobacillus brevis]). The monomers of these amino acid decarboxylases all consist of three characteristic domains (Figure 3 and Supplementary figure 4). The C-terminal domain, also known as the small domain, is composed of three antiparallel β-sheets and three α-helices, facing toward the opposite monomers. The large domain contains seven stranded β-sheets surrounded by eight α-helices, exhibiting characteristics typical of Type I enzymes. This domain includes a conserved PLP-binding active center and amino acid residues that are conserved across species. The N-terminal domain consists of elongated α-helices and is connected to the large domain via a loop. The helix of one subunit is antiparallel to the corresponding helix in the neighboring subunit, forming a clamp. Therefore, the N-terminal domain may not be an independent folding unit and is likely only stable in dimers.

The monomeric structure of CsAlaDC and AtSerDC exhibits numerous structural features common to other Group II PLP-dependent amino acid decarboxylases, and the amino acid residues that bind to PLP cofactors are conserved across multiple enzymes (Supplementary figure 3 and 4). Out of the eight proteins analyzed, a total of thirteen amino acid residues were found to be conserved across all sequences (Glu143, Glu171, Lys201, Gly245, Thr247, Asp253, His275, Asp277, Ala279, Pro286, Ser302, Lys309, and Tyr336 in CsAlaDC). Some of them are situated within the active center and play a role in stabilizing PLP or facilitating catalytic reactions. Conversely, other residues reside outside the active center and their function remains unclear. (Supplementary figure 3). PLP is bound to conserved lysine residues in the active center through the Schiff base. The carboxylic group of conserved Asp residues stabilizes protonated pyridine nitrogen of PLP by electrostatic action, providing the latter with strong electrophilicity to stabilize carbanion intermediates during enzymatic catalysis. Another conserved His residue is situated laterally to the PLP, while on the opposite side, there is a conserved Ala residue. The oxygen atom at position 3 of PLP is stabilized by hydrogen bonding with the Thr residue. The phosphate moiety of PLP is additionally secured to the protein through an elongated system of hydrogen bonding. The α5 helix located within the large domain exhibits evolutionary conservation in Group II enzymes, facilitating interaction with the PLP phosphate moiety and conferring appropriate stabilization thereof.

The superimposition of the crystal structures of CsAlaDC and AtSerDC with those of related enzymes suggests that residue Phe106 in CsAlaDC (corresponding to Tyr in AtSerDC and Glu in MetDC) may play a role in determining substrate specificity. This hypothesis was tested via mutagenesis experiments, which revealed that substitution of Phe106 for Tyr rendered CsAlaDC inactive, while substitution of Tyr111 for Phe in AtSerDC enabled alanine decarboxylase activity. Likewise, the substitution of Glu64 for Ala in MetDC reduced enzyme activity by approximately 50-fold compared to the wild-type (35), supporting this conclusion.

Group II amino acid decarboxylase is characterized by the existence of highly flexible catalytic rings, which are of great significance for the catalytic mechanism of decarboxylase (31). The catalytic rings are located at the dimer interface and extend to the active sites of other monomers in a closed conformation. In the structure of CsAlaDC, the catalytic ring is orderly and allocated, while in the structure of AtSerDC, it is disordered and lacks electron density (Figure 3F). Prior research has established that the conserved amino acid residue Tyr within the ring plays a crucial role in catalysis by donating protons to carbanions of quinone intermediates that arise following decarboxylation (36, 37). Our experiments demonstrate that the substitution of the corresponding Tyr with Phe in the loop renders the protein inactive. Furthermore, alterations in the structure of CsAlaDC when combined with EA provide further evidence supporting this conclusion. Specifically, the structure of CsAlaDC-EA complex displayed a significant conformational change in Tyr336* located in the catalytic site (as depicted in Figure 3G). Comparing the structure of CsAlaDC apo, the CsAlaDC-EA complex exhibited a notable 60-degree shift in the p-hydroxyl group of Tyr336* along the Cα-Cβ bond.

The monomeric configuration of CsAlaDC and AtSerDC is akin to that of other Group II amino acid decarboxylases, except for a conspicuous dissimilarity. Unlike other Group II amino acid decarboxylases, the C-terminal of CsAlaDC and AtSerDC harbors an evident zinc finger structure. We propose that this structure could potentially influence enzyme stability. To test this hypothesis, we truncated the zinc finger structure from both proteins and expressed them. Our findings indicate that, following the excision of the zinc finger structure, CsAlaDC became insoluble, while AtSerDC displayed a similar trait to a certain extent.

Evolution of AlaDC and SerDC

The sequence identity between CsAlaDC and AtSerDC is 74.5%, indicating a significant degree of homology. Moreover, the two proteins exhibit a high level of monomer structural similarity. Notably, they differ in substrate specificity. Using the crystal structure of CsAlaDC and AtSerDC as a basis, we identified pivotal amino acid residues that govern the substrate specificity of serine or alanine decarboxylase. We observed that sequences proximal to these sites exhibit a high degree of conservation in both proteins and possess characteristic motifs (Figure 4B). Moreover, we analyzed serine decarboxylase-like proteins in Embryophyta that are highly homologous to CsAlaDC and constructed phylogenetic trees to gain insight into their evolutionary relationships. Based on our experimental findings and evolutionary evidence, we have identified a conserved YPX motif at the substrate binding pocket. For most plants, the residue X in this motif is predominantly Tyr, which corresponds to the serine decarboxylase protein. However, other residues such as Thr, Phe, Val, Ala, and Ile can also occupy this position (Figure 4B and C). Interestingly, serine decarboxylase-like proteins containing YPF motifs, which have the potential to function as alanine decarboxylases, were found to be distributed throughout the phylogenetic tree, including Asteroideae, Ericales, Chenopodiaceae, Poaceae, and Ranunculales (Supplementary figure 1B). However, these proteins were absent in some more recent species. This observation implies that the emergence of alanine decarboxylase in these particular species could be attributed to convergent evolution.

Moreover, we have observed the enrichment of serine decarboxylase-like proteins that possess a YPT motif within the Fabales. Prior research has demonstrated that XP_004496485-1, which possesses a YPT motif, displays catalytic efficacy toward the processes of decarboxylation and oxidative deamination of Phe, Met, Leu, and Trp (38). Given that the YPT motif is highly conserved and widely distributed in Fabales, serine decarboxylase-like proteins bearing the YPT motif may have developed a unique substrate specificity in Fabales, beyond their conventional decarboxylation functions, as exemplified by XP_004496485-1 protein mentioned above. We speculate that these proteins may be capable of catalyzing other reactions, potentially involving non-protein amino acids or other substances as substrates.

Applied to improve the synthesis of theanine

Theanine is an important indicator of green tea quality. Therefore, improving the synthesis of theanine in tea plants is the focus of research. In this study, through crystal structure analysis and mutation verification, we found that the catalytic activity of CsAlaDCL110F/P114A is 2.3 times higher than that of the wild-type protein, resulting in a more abundant synthesis of theanine in vitro. This gives us great inspiration to improve the synthesis of EA in tea plants by gene editing, thus increasing the content of theanine in tea plants (Figure 5).

Theanine is highly demanded, by the market, due to its health effects and medicinal value, and as a constituent in food, cosmetics and in other fields. To meet the market demand, a variety of methods have been used to acquire theanine, with the main methods including direct extraction, chemical synthesis, biotransformation (microbial fermentation) and plant cell culture. Based on this study, we can use mutation technology to modify engineering bacteria and improve the possibility of synthesis of theanine in bacteria, yeast, model plants, etc, promoting its use for industrial theanine production.

Conclusions

In summary, we determined the crystal structures of the CsAlaDC-EA complex and AtSerDC. We employed structural biology to reveal the substrate selection mechanism of these two decarboxylases. By structural analysis and enzyme activity experiments, we found that Tyr111 of AtSerDC and Phe106 of CsAlaDC as crucial sites for substrate specificity. Next, through mutation screening, we identified a CsAlaDC mutant protein (L110F/P114A) with catalytic activity 2.3 times higher than that of the wild-type protein. In an in vitro L-theanine biosynthesis system with added PsGS or GMAS, we demonstrated that the L-theanine yield of the CsAlaDCL110F/P114A group was 4.3 and 6.7 times higher than that of the wild-type control group, respectively. In addition, by utilizing the substrate recognition key motif of CsAlaDC and AtSerDC, evolutionary analysis was conducted on protein sequences exhibiting high homologous to CsAlaDC in Embryophyta, which resulted in the discovery of 13 potential alanine decarboxylases. Furthermore, we discovered serine protease-like proteins containing special motifs that have evolved in Fabale, which may possess unique substrate specificity and catalytic functions. On the other hand, our study has effectively enhanced the efficiency of the initial stage of theanine synthesis and is anticipated to offer a promising avenue for developing a novel approach to producing L-theanine, thereby facilitating the progress of the tea industry.

Materials and Methods

Plants materials

Tobacco (Nicotiana benthamiana) plants were grown in a controlled chamber, under a 16-h light and 8-h dark photoperiod at 25 ℃. Leaves of 5-week-old tobacco plants were used for transient transformation, mediated by Agrobacterium tumefaciens strain GV3101.

Gene cloning and protein expression

The coding sequences of AtSerDC and CsAlaDC, amino acid residues 66-482 and 61-478, respectively, were amplified using PCR and then ligated into the Nde Ι and Xho I restriction sites of the pET-28a and pET-22b vector, respectively, providing the recombinant vector pET-28a-AtSerDC and pET-22b-CsAlaDC. The cDNAs encoding CsSerDC were amplified using PCR and then ligated into the Nde Ι and Xho I restriction sites of the pET-28a vector. Gene-specific primers were listed in Supplementary table 2. The recombinant plasmid was transformed into E. coli BL21 (DE3) competent cells. Positive transformants were grown in a 5 mL LB medium containing 30 μg/mL kanamycin or 50 μg/mL ampicillin at 37 °C overnight and then subcultured into an 800 mL LB medium containing the corresponding antibiotic. Protein expression was induced by the addition of 0.2 mM isopropyl-β-d-thiogalactoside (IPTG) for 20 h at 16 ℃ when the optical density (OD) at 600 nm reached 0.6-0.8, harvested by centrifugation at 4 °C and 4,000 rpm for 30 min.

Protein production and crystallization

The cell pellet was suspended in 30 mL of lysis buffer (20 mM HEPES, pH 7.5, 200 mM NaCl, 0.1 mM PLP), disrupted by High-Pressure Homogenizer, and then centrifuged at 16,000 rpm for 30 min at 4 °C to remove the cell debris. The supernatants were purified with a Ni-Agarose resin column followed by size-exclusion chromatography. Before crystallization, purified proteins were concentrated at 10 mg/mL. Crystallization conditions were screened by the sitting-drop vapor diffusion method using the reservoir solutions supplied in commercially available screening kits (Crystal Screen, Crystal Screen 2, PEGRx 1, 2, and SaltRx 1, 2). A droplet made by mixing 1.0 μL of purified AtSerDC or CsAlaDC (10 mg/mL) with an equal volume of a reservoir solution was equilibrated against 100 μL of the reservoir solution at 16℃. The crystal of AtSerDC was obtained using buffer pH 8.0 containing 20% (w/v) PEG400 as a precipitate and 0.2 M CaCl2. The crystal of the CsAlaDC-EA complex was obtained at pH 7.5 containing 2.6 M sodium acetate. The crystal of CsAlaDC was obtained at pH 6.0 containing 3.5 M sodium formate.

Data collection and processing

Crystals were grown by sitting-drop vapor diffusion method at 16 °C. The volume of the reservoir solution was 100 µL and the drop volume was 2 µL, containing 1 µL of protein sample and 1 µL of reservoir solution. The reservoir solution of AtSerDC contained 0.1 M Tris-HCl (pH 8.0), 0.2 M CaCl2, and 20% PEG 400. The reservoir solution of CsAlaDC contained 0.1 M HEPES (pH 7.5), 2.6 M sodium acetate or 0.1 M Bis-Tris (pH 6.0), and 3.5 M sodium formate. Crystals grew in 3-5 days using a protein concentration of 10 mg/mL. Diffraction data were collected at Shanghai Synchrotron Radiation Facility (China). The collected data sets were indexed, integrated, and scaled using the HKL3000 software package. The structure of AtSerDC was solved by molecular replacement using the structure of HisDC (PDB code: 7ERV [Photobacterium phosphoreum]) as the model, and utilizing AtSerDC as a molecular displacement model for both CsAlaDC and the CsAlaDC-EA complex. The AtSerDC, CsAlaDC, and CsAlaDC-EA complex exhibit resolutions of 2.85 Å, 2.50 Å, and 2.60 Å, respectively. The statistics for data collection and processing are summarized in Supplementary table 1.

Enzyme activity assays

Decarboxylase activity was measured by detecting products (EA or ethanolamine) in Waters Acquit ultraperformance liquid chromatography (UPLC) system (20, 39). The 100 μL reaction mixture, containing 20 mM substrate (Ala or Ser), 100 mM potassium phosphate, 0.1 mM PLP, and 0.025 mM purified enzyme, was prepared and incubated at standard conditions (45 °C and pH 8.0 for CsAlaDC, 40 °C and pH 8.0 for AtSerDC for 30 min). Then, the reaction was stopped with 20 μL of 10% trichloroacetic acid.

The detection methodology for theanine production remains consistent with the aforementioned approach. The 100 μL reaction mixture, containing 20 mM Ala, 45 mM Glu, 50 mM HEPES (pH 7.5), 0.1 mM PLP, 30 mM MgCl2, 10 mM ATP, 0.03 mM PsGS/GMAS, and 0.025 mM CsAlaDC/CsAlaDC L110F/CsAlaDC L110F/P114A, was prepared and incubated at standard conditions for 1 h. Subsequently, the reaction was terminated via immersion of the reaction vessel in a metal bath at 96 °C for 3 min.

The product was derivatized with 6-aminoquinolyl-N-hydroxy-succinimidyl carbamate (AQC) and subjected to analysis by UPLC. All enzymatic assays were performed in triplicate.

Site-directed mutagenesis

Site-directed mutagenesis experiment was conducted using a PCR method from the wild-type construct pET-28a-AtSerDC and pET-22b-CsAlaDC, respectively. Dpn I endonucleases were used to digest the parental DNA template. The reaction mixture was used to transform E. coli DH5α competent cells and the plasmids from positive strains were extracted to E. coli BL21 (DE3) for protein expression, purification, and analyzing the enzymatic activity.

In vivo enzyme activity assay in N. benthamiana

The amplified PCR products were fused to the plant expression vector, pCAMBIA1305. Linearization was conducted by restriction digest with Spe I and BamH I. The recombinant colonies were selected for PCR validation on the appropriate antibiotics plate. After validation, the plasmids were electroporated into Agrobacterium tumefaciens strain GV3101(pSoup-p19). Empty pCAMBIA1305 vector, with an intron containing the GFP gene, was treated as the control.

Agrobacterium transient expression assays were performed on 5-week-old N. benthamiana plants. Agrobacterium tumefaciens strain GV3101 (pSoup-p19), carrying the above-described vectors, were cultured in Luria Bertani (LB) medium, containing appropriate antibiotics, at 28 °C. When the absorbance of bacteria colonies reached OD600 = 0.6-0.8, bacterial cells were collected and resuspended in MMA solution (10 mM MgCl2, 10 mM 2-(N-morpholino) ethane sulfonic acid (MES), pH 5.6). After the OD600 of the resuspended bacterial solution was adjusted to approx. 1.0, acetosyringone (AS) was added with a final concentration of 200 µM, and this solution was then incubated, at room temperature, for at least 3 h in darkness. Next, cell suspensions were infiltrated into N. benthamiana leaves with a needle-free syringe. These N. benthamiana leaves were then collected 3 days post infiltration, frozen in liquid nitrogen and stored at −80 °C. Internal EA in N. benthamiana leaves was extracted as previously described (22), and then the solvent was subjected to gas chromatography-mass spectrometry GS-MS system.

Transcript level analysis in N. benthamiana

Total RNA was isolated from samples using the RNAprep Pure Plant Kit (Tiangen, Beijing, China), according to the manufacturer’s protocol. The cDNAs were synthesized using TransScript One-Step gDNA Removal and cDNA Synthesis SuperMix Kit (TransGen Biotech, Beijing, China). The qRT-PCR assays were performed, as previously described (40). Primers used for qRT-PCR assays were listed in Supplementary Table 3 and the qRT-PCR was run on a Bio-Rad CFX96TM RT PCR detection system and CFX Manager Software. Each reaction reagent (20 μL) contained 0.4 μL forward and reverse primers (10 μM), 2 μL cDNA (200±5 ng/μL), 10 μL SYBR Green Supermix (Vazyme, Nanjing, China) and 6.2 μL double-distilled water. Reaction was performed by a two-step method: 95 ℃ for 5 min; 40 cycles of 95 ℃ for 10 s; and 60 ℃ for 30 s. The glyceraldehyde-3-phosphate dehydrogenase (GAPDH) gene was used for internal normalization in each RT-PCR, and the 2−ΔΔCT method (41) was used to calculate the relative gene expression. All samples were performed with three replicates.

Multiple sequence alignment

MUSCLE was used to generate the protein multiple sequence alignment of amino acid decarboxylases with default settings (42). ESPript 3.x was used to display the multiple sequence alignment (43).

Phylogenetic analysis

The CsAlaDC was used as a query for searching its homologs in Embryophyta using BLASTP (44) (version 2.13.0). To fetch homologs of CsAlaDC in distant species, we performed a two-round blast. We first search for the best hit in the given order under Embryophyta. Then we blast each species under a provided order using the order best hit as a query and as many as 10 hits were kept. The sequences with e values <= 0.001 were removed. We further dropped sequences with lengths less than 200 or larger than 800 amino acids. Then multiple sequence alignment was performed via clustalo (version 1.2.4) with default settings on these filtered sequences (45). The NJ algorithm implemented in clustalw (version 2.1) was adopted to build the phylogenetic tree (46). The R package ggtree (version 3.2.1) and universalmotif (version 1.12.4) were used for tree and motif visualization (47).

eLife Assessment

This study reports biochemical and structural analysis of two PLP decarboxylase enzymes from plants. The findings are useful due to the utility of these enzymes in industrial theanine production. While certain aspects of the study are solid, other components elucidating the role of a Zn(II)-binding motif are incomplete. In addition, some of the finding could be presented more clearly, including the connections between the structural findings and the reaction mechanism. The work will be of interest to enzymologists studying PLP enzymes and those interested in enzyme engineering in plants.

Reviewer #1 (Public Review):

In this study, the structural characteristics of plant AlaDC and SerDC were analyzed to understand the mechanism of functional differentiation, deepen the understanding of substrate specificity and catalytic activity evolution, and explore effective ways to improve the initial efficiency of theanine synthesis.

On the basis of previous solid work, the authors successfully obtained the X-ray crystal structures of the precursors of theanine synthesis-CsAlaDC and AtSerDC, which are key proteins related to ethylamine synthesis, and found a unique zinc finger structure on these two crystal structures that are not found in other Group II PLP- dependent amino acid decarboxylases. Through a series of experiments, it is pointed out that this characteristic zinc finger motif may be the key to the folding of CsAlaDC and AtSerDC proteins, and this discovery is novel and prospective in the study of theine synthesis.

In addition, the authors identified Phe106 of CsAlaDC and Tyr111 of AtSerDC as key sites of substrate specificity by comparing substrate binding regions and identified amino acids that inhibit catalytic activity through mutation screening based on protein structure. It was found that the catalytic activity of CsAlaDCL110F/P114A was 2.3 times higher than that of CsAlaDC. At the same time, CsAlaDC and AtSerDC substrate recognition key motifs were used to carry out evolutionary analysis of the protein sequences that are highly homologous to CsAlaDC in embryos, and 13 potential alanine decarboxylases were found, which laid a solid foundation for subsequent studies related to theanine synthesis.

In general, this study has a solid foundation, the whole research idea is clear, the experimental design is reasonable, and the experimental results provide strong evidence for the author's point of view. Through a large number of experiments, the key links in the theanine synthesis pathway are deeply studied, and an effective way to improve the initial efficiency of theanine synthesis is found, and the molecular mechanism of this way is expounded. The whole study has good novelty and prospectivity, and sheds light on a new direction for the efficient industrial synthesis of theanine.

Reviewer #2 (Public Review):

Summary:

The manuscript focuses on the comparison of two PLP-dependent enzyme classes that perform amino acyl decarboxylations. The goal of the work is to understand the substrate specificity and factors that influence the catalytic rate in an enzyme linked to theanine production in tea plants.

Strengths:

The work includes x-ray crystal structures of modest resolution of the enzymes of interest. These structures provide the basis for the design of mutagenesis experiments to test hypotheses about substrate specificity and the factors that control catalytic rate. These ideas are tested via mutagenesis and activity assays, in some cases both in vitro and in plants.

Weaknesses:

The manuscript could be more clear in explaining the contents of the x-ray structures and how the complexes studied relate to the reactant and product complexes. The structure and mechanism section would also be strengthened by including a diagram of the reaction mechanism and including context about reactivity. As it stands, much of the structural results section consists of lists of amino acids interacting with certain ligands without any explanation of why these interactions are important or the role they play in catalysis. The experiments testing the function of a novel Zn(II)-binding domain also have serious flaws. I don't think anything can be said at this point about the function of the Zn(II) due to a lack of key controls and problems with experimental design.

Reviewer #3 (Public Review):

In the manuscript titled "Structure and Evolution of Alanine/Serine Decarboxylases and the Engineering of Theanine Production," Wang et al. solved and compared the crystal structures of Alanine Decarboxylase (AlaDC) from Camellia sinensis and Serine Decarboxylase (SerDC) from Arabidopsis thaliana. Based on this structural information, the authors conducted both in vitro and in vivo functional studies to compare enzyme activities using site-directed mutagenesis and subsequent evolutionary analyses. This research has the potential to enhance our understanding of amino acid decarboxylase evolution and the biosynthetic pathway of the plant-specialized metabolite theanine, as well as to further its potential applications in the tea industry.

Author Response

The following is the authors’ response to the original reviews.

Response to reviewer’s comments

Reviewer #1 (Public Review):

In this study, the structural characteristics of plant AlaDC and SerDC were analyzed to understand the mechanism of functional differentiation, deepen the understanding of substrate specificity and catalytic activity evolution, and explore effective ways to improve the initial efficiency of theanine synthesis.

On the basis of previous solid work, the authors successfully obtained the X-ray crystal structures of the precursors of theanine synthesis-CsAlaDC and AtSerDC, which are key proteins related to ethylamine synthesis, and found a unique zinc finger structure on these two crystal structures that are not found in other Group II PLP-dependent amino acid decarboxylases. Through a series of experiments, it is pointed out that this characteristic zinc finger motif may be the key to the folding of CsAlaDC and AtSerDC proteins, and this discovery is novel and prospective in the study of theine synthesis.

In addition, the authors identified Phe106 of CsAlaDC and Tyr111 of AtSerDC as key sites of substrate specificity by comparing substrate binding regions and identified amino acids that inhibit catalytic activity through mutation screening based on protein structure. It was found that the catalytic activity of CsAlaDCL110F/P114A was 2.3 times higher than that of CsAlaDC. At the same time, CsAlaDC and AtSerDC substrate recognition key motifs were used to carry out evolutionary analysis of the protein sequences that are highly homologous to CsAlaDC in embryos, and 13 potential alanine decarboxylases were found, which laid a solid foundation for subsequent studies related to theanine synthesis.

In general, this study has a solid foundation, the whole research idea is clear, the experimental design is reasonable, and the experimental results provide strong evidence for the author's point of view. Through a large number of experiments, the key links in the theanine synthesis pathway are deeply studied, and an effective way to improve the initial efficiency of theanine synthesis is found, and the molecular mechanism of this way is expounded. The whole study has good novelty and prospectivity, and sheds light on a new direction for the efficient industrial synthesis of theanine

Response: Thank you very much for taking time to review this manuscript. We appreciate all your insightful comments and constructive suggestions.

Reviewer #1 (Recommendations For The Authors):

(1) If some test methods are not original, references or method basis should be indicated.

Response: Thank you very much for your careful reading of the manuscript and valuable suggestions. We have added references for the enzymatic activity experiments performed to measure the synthesis of theanine in the revised manuscript.

(2) The conclusion is a little lengthy, and the summary of the whole study is not well condensed.

Response: Thank you very much for your valuable suggestions. We have refined the conclusion in the revised manuscript, and it is as follows:

In conclusion, our structural and functional analyses have significantly advanced understanding of the substrate-specific activities of alanine and serine decarboxylases, typified by CsAlaDC and AtSerDC. Critical amino acid residues responsible for substrate selection were identified—Tyr111 in AtSerDC and Phe106 in CsAlaDC—highlighting pivotal roles in enzyme specificity. The engineered CsAlaDC mutant (L110F/P114A) not only displayed enhanced catalytic efficiency but also substantially improved L-theanine yield in a synthetic biosynthesis setup with PsGS or GMAS. Our research expanded the repertoire of potential alanine decarboxylases through the discovery of 13 homologous enzyme candidates across embryophytic species and uncovered a special motif present in serine protease-like proteins within Fabale, suggesting a potential divergence in substrate specificity and catalytic functions. These insights lay the groundwork for the development of industrial biocatalytic processes, promising to elevate the production of L-theanine and supporting innovation within the tea industry.

Reviewer #2 (Public Review)

Summary:

The manuscript focuses on the comparison of two PLP-dependent enzyme classes that perform amino acyl decarboxylations. The goal of the work is to understand the substrate specificity and factors that influence the catalytic rate in an enzyme linked to theanine production in tea plants.

Strengths:

The work includes x-ray crystal structures of modest resolution of the enzymes of interest. These structures provide the basis for the design of mutagenesis experiments to test hypotheses about substrate specificity and the factors that control catalytic rate. These ideas are tested via mutagenesis and activity assays, in some cases both in vitro and in plants.

Weaknesses:

The manuscript could be more clear in explaining the contents of the x-ray structures and how the complexes studied relate to the reactant and product complexes. The structure and mechanism section would also be strengthened by including a diagram of the reaction mechanism and including context about reactivity. As it stands, much of the structural results section consists of lists of amino acids interacting with certain ligands without any explanation of why these interactions are important or the role they play in catalysis. The experiments testing the function of a novel Zn(II)-binding domain also have serious flaws. I don't think anything can be said at this point about the function of the Zn(II) due to a lack of key controls and problems with experimental design.

Response: Thank you very much for your thoughtful comments and feedback on our manuscript. We are pleased to hear that the work's strengths, such as the X-ray crystal structures and the mutagenesis experiments tied to the catalytic rate and substrate specificity, align with the goals of our research.

We recognize the areas identified for improvement and appreciate the suggestions provided. We have emphasized how we use the structural information obtained to infer the roles of key amino acid residues in the reaction. Additionally, we have added a diagram of the reaction mechanism in the Supplementary figure to provide clearer context on reactivity and improve the overall understanding of the catalytic process. Regarding the structural results section, we have included a discussion that contextualizes the list of amino acids and their interactions with the ligands by explaining their significance and roles in catalysis. We acknowledge the weaknesses you've pointed out in the experiments concerning the novel Zn(II)-binding domain, but we would like to clarify that the focus of our study was not primarily on the zinc structure. While we agree that there may be limitations in the experimental design and controls for the zinc binding domain, we believe that these flaws do not significantly impact the overall findings of the study. The experiment served as a preliminary exploration of the potential functionality of the domain, and further studies are required to fully understand its role and mechanism.

Reviewer #2 (Recommendations For The Authors):

(1) In addition to the points raised in the public review, it would be ideal to provide some context for the enzymatic characterization. Why are the differences in kinetic parameters for AlaDC and SerDC significant?

Response: Thank you for your comments and suggestions. The Km values for CsAlaDC and SerDCs are comparable, suggesting similar substrate affinities. However, CsAlaDC exhibits a significantly lower Vmax compared to AtSerDC and CsSerDC. This discrepancy implies that CsAlaDC and SerDCs may differ in the rates at which they convert substrate to product when saturated with substrate. SerDCs may have a faster turnover rate, meaning they convert substrate to product and release the enzyme more quickly, resulting in a higher Vmax. Differences in the stability or correct folding of the enzymes under assay conditions can also affect their Vmax. If SerDCs are more stable, they might maintain their catalytic activity better at higher substrate concentrations, contributing to a higher Vmax. We have added these to the part of “Enzymatic properties of CsAlaDC, AtSerDC, and CsSerDC” in our revised manuscript.

(2) Why is Phe106/Tyr111 pair critical for substrate specificity? Does the amino acid contact the side chain? It might be helpful to a reader to formulate a hypothesis for this interaction.

Response: Thank you for the question and comments. We conducted a comparison between the active sites of CsAlaDC and AtSerDC and observed a distinct difference in only two amino acids: F106 in CsAlaDC and Y111 in AtSerDC. The remaining amino acids were found to be identical. Expanding on previous research concerning Group II PLP-dependent amino acid decarboxylases, it was postulated and subsequently confirmed that these specific amino acids play a crucial role in substrate recognition. However, since we lack the structure of the enzyme-substrate complex, we are unable to elucidate the precise interactions occurring between the substrate and the amino acids at this particular site based solely on structural information.

(3) Line 55 - Define EA again.

Response: Thank you very much for your careful reading of the manuscript and valuable suggestions. We have redefined “EA” as the abbreviation for ethylamine in the revised manuscript.

(4) Line 58 - The meaning of "determined by the quality formation of tea" is not clear.

Response: Thank you very much for your careful reading of the manuscript and valuable suggestions. We have modified it in the revised manuscript.

(5) Line 65 - Missing words between "despite they".

Response: Thank you very much for your careful reading of the manuscript. We have corrected it in the revised manuscript.

(6) Line 67 - Need a reference for the statement about lower activity?

Response: Thank you for the question and comments. We have provided the following reference to support this statement in the revised manuscript.

Reference: Bai, P. et al. (2021) Biochemical characterization of specific Alanine Decarboxylase (ADC) and its ancestral enzyme Serine Decarboxylase (SDC) in tea plants (Camellia sinensis). BMC Biotechnol. 21,17.

(7) Line 100-101 - The meaning of "its closer relationship was Dicots plants." is not clear.

Response: We have revised the sentence in the revised manuscript, as follows: “Phylogenetic analysis indicated that CsAlaDC is homologous with SerDCs in Dicots plants.”

(8) Line 139 - Missing a word between "as well as" and "of".

Response: Thank you very much for your careful reading of the manuscript and valuable suggestions. We have corrected it in the revised manuscript.

(9) Line 142 - The usage of comprised here is not correct. It would be more correct to say "The overall architecture of CsAlaDC and AtSerDC is homodimeric with the two subunits...".

Response: Thank you very much for your careful reading of the manuscript and valuable suggestions. We have corrected it in the revised manuscript.

(10) Line 148-149 - I didn't understand the statement about the "N-terminal structures" Are these structures obtained from protein samples that have a truncated N-terminus?

Response: Group II PLP-dependent amino acid decarboxylases are comprised of three distinct structural domains: the N-terminal domain, the large domain, and the C-terminal domain. Each of these domains possesses unique structural features. Similarly, CsAlaDC and AtSerDC can also be classified into three structural domains based on their specific characteristics. To achieve more stable proteins for further experiments, we conducted truncation on both of these proteins. The truncated section pertains to a subsection of the N-terminal domain and is truncated from the protein's N-terminus.

(11) Line 153 - Say "is composed of" instead of "composes of".

Response: Thank you very much for your careful reading of the manuscript and valuable suggestions. We have corrected it in the revised manuscript.

(12) Line 156 - I didn't understand the statement about the cofactor binding process. What is the cofactor observed? And how can we say anything about the binding process from a single static structure of the enzyme? It might be better to say that the cofactor binding site is located at the subunit junction - but the identity of the cofactor still needs to be defined first.

Response: Thank you for your comments and suggestions. The cofactor mentioned here is PLP. We aim to elucidate the binding state of PLP at the active site, excluding the binding process. The description has been revised in the revised manuscript.

(13) Lines 157-158 - I didn't understand the conclusion about the roles of each monomer. In the images in Figure 3 - both monomers appear to bind PLP but the substrate is not present - so it's not clear how conclusions can be drawn about differential substrate binding in the two subunits.

Response: Thank you very much for your careful reading and valuable suggestions. The main idea we want to convey is that this protein possesses two active sites. At each active site, the two monomers carry out distinct functions. Of course, our previous conclusion is inaccurate due to the non-existence of the substrate. So, we have made the necessary amendments in the revised manuscript.

(14) Line 161 - I would say loop instead of ring.

Response: Thank you very much for your careful reading of the manuscript and valuable suggestions. We have corrected it in the revised manuscript.

(15) Line 165 - Please provide some references for this statement. It would also be ideal to state the proximity of the Zn-binding motif to the active site or otherwise provide some information about the role of the motif based on its location.

Response: Thank you for your comments and suggestions. We have provided the following references to support this statement in the revised manuscript.

(A) Structure of histidine decarboxylase.

(B) Structure of glutamate decarboxylase.

Reference:

30 Komori, H. et al. (2012) Structural study reveals that Ser-354 determines substrate specificity on human Histidine Decarboxylase. J Biol Chem. 287, 29175-83.

31 Huang, J. et al. (2018) Lactobacillus brevis CGMCC 1306 glutamate decarboxylase: Crystal structure and functional analysis. Biochem Biophys Res Co. 503, 1703-1709

In CsAlaDC, the zinc is positioned at a distance of 29.6 Å from the active center, whereas in AtSerDC, the zinc is situated 29 Å away from the active center. Hence, we hypothesize that this structure does not impact the enzyme's catalytic activity but might be correlated with its stability.

(16) Lines 166-178 - This paragraph appears to be a list of all of the interactions between the protein, PLP, and the EA product. It would be ideal to provide some text to explain why these interactions are important and what we can learn from them.

Response: Thank you very much for your careful reading of the manuscript and valuable suggestions. We have been conducting additional analysis on the functional roles of amino acid residues involved in the interaction between the active site and PLP. This analysis focuses on aiding PLP binding, determining its orientation, and understanding enzyme catalytic mechanisms. These details are mentioned in the revised manuscript.

(17) Line 192 - Bond not bound.

Response: Thank you very much for your careful reading of the manuscript and valuable suggestions. We have made corrections in the revised manuscript.

(18) Lines 201-207 - It would be ideal to verify that the inclusion of 5 mM DTT affects Zn binding. It's not clear to me that this reagent would necessarily disrupt Zn binding. Under certain circumstances, it could instead promote Zn association. For example, if the Cys ligands are oxidized initially but then become reduced? I don't think the current experiment really provides any insight into the role of the Zn.

Response: Thank you for your valuable insights regarding the role of DTT and its potential effects on Zn binding in our experiments. The main function of DTT is to protect or restore the reduced state of proteins and other biological molecules, particularly by disrupting the crosslinking formed by thiol (-SH) groups and disulfide bonds to maintain the function and structure of proteins. Therefore, the reason for DTT's inhibition of enzyme activity is unknown, and we cannot provide a reasonable explanation for this phenomenon. As a result, we have removed the section discussing the inhibition of enzyme activity by DTT in our revised manuscript.

Reviewer #3 (Public Review):

In the manuscript titled "Structure and Evolution of Alanine/Serine Decarboxylases and the Engineering of Theanine Production," Wang et al. solved and compared the crystal structures of Alanine Decarboxylase (AlaDC) from Camellia sinensis and Serine Decarboxylase (SerDC) from Arabidopsis thaliana. Based on this structural information, the authors conducted both in vitro and in vivo functional studies to compare enzyme activities using site-directed mutagenesis and subsequent evolutionary analyses. This research has the potential to enhance our understanding of amino acid decarboxylase evolution and the biosynthetic pathway of the plant-specialized metabolite theanine, as well as to further its potential applications in the tea industry. Response: Thank you very much for taking the time to review this manuscript. We appreciate all your insightful comments.

Reviewer #3 (Recommendations For The Authors):

Page 6, Figure 2, Page 23 (Methods)

"The supernatants were purified with a Ni-Agarose resin column followed by size-exclusion chromatography."

What kind of SEC column did the authors use? Can the authors provide the SEC elution profile comparison results and size standard curve?

Response: We use a Superdex 200 (Hiload 16/600) column for size exclusion chromatography. The comparison results of SEC elution profiles for AtSerDC and CsAlaDC, along with the standard curve of SEC column, are presented below.

(A) Comparison of elution profiles of CsAlaDC and AtSerDC.

(B) Elution profile of Blue Dextron 2000. (C) Elution profile of mixed protein (Aldolase, 158000 Da,71.765ml; Conalbumin, 75000 Da,79.391ml; Ovalbumin, 44000 Da,83.767ml; Carbonic anhydrase, 29000 Da,90.019ml; Ribonuclease A, 13700 Da,98.145ml). (D) Size standard curves of Superdex 200 (Hiload 16/600) column.

Page 6 & Page 24 (Methods)

"The 100 μL reaction mixture, containing 20 mM substrate (Ala or Ser), 100 mM potassium phosphate, 0.1 mM PLP, and 0.025 mM purified enzyme, was prepared and incubated at standard conditions (45 ℃ and pH 8.0 for CsAlaDC, 40 ℃ and pH 8.0 for AtSerDC for 30 min)."

(1) The enzymatic activities of CsAldDC and AtSerDC were measured at two different temperatures (45 and 40 ℃, but their activities were directly compared. Is there a reason for experimenting at different temperatures?

Response: We determined that the optimal reaction temperature for AtSerDC is 40°C and for CsAlaDC is 45°C through our verification process. Consequently, all subsequent experiments were performed at these specific temperatures.

(A) Relative activity of CsAlaDC at different temperatures.

(B) Relative activity of AtSerDC at different temperatures.

(2) Enzyme activities were measured at temperatures above 40℃, which is not a physiologically relevant temperature and may affect the stability or activity of the proteins. At the very least, the authors should provide temperature-dependent protein stability data (e.g., CD spectra analysis) or, if possible, temperature-dependent enzyme activities, to show that their experimental conditions are suitable for studying the activities of these enzymes.

Response: Thank you very much for your careful reading. We have already validated that the experimental temperature we used did not significantly affect the stability of the protein before experimenting. The results are shown in the figure below:

Place the two proteins individually into water baths set at temperatures of 25°C, 37°C, 45°C, 60°C, and 80°C for 15 minutes. Subsequently, carry out enzymatic reactions utilizing a standard reaction system, with untreated enzymes serving as the experimental control within the said system. The experimental results suggest that the temperature at which we experimented does not have a significant impact on the stability of the enzyme.

(3) The authors used 20 mM of substrate. What are the physiological concentrations of alanine and serine typically found in plants?

Response: The content of alanine in tea plant roots ranges from 0.28 to 4.18 mg/g DW (Yu et al., 2021; Cheng et al., 2017). Correspondingly, the physiological concentration of alanine is 3.14 mM to 46.92 mM, in tea plant roots. The content of serine in plants ranges from 0.014 to 17.6 mg/g DW (Kumar et al., 2017). Correspondingly, the physiological concentration of serine is 0.13 mM to 167.48 mM in plants. In this study, the substrate concentration of 20 mM was close to the actual concentrations of alanine and serine in plants.

Yu, Y. et al. (2021) Glutamine synthetases play a vital role in high accumulation of theanine in tender shoots of albino tea germplasm "Huabai 1". J. Agric. Food Chem. 69 (46),13904-13915.

Cheng, S. et al. (2017) Studies on the biochemical formation pathway of the amino acid L-theanine in tea (Camellia sinensis) and other plants.” J. Agric. Food Chem. 65 (33), 7210-7216.

Kumar, V. et al. (2017) Differential distribution of amino acids in plants. Amino Acids. 49(5), 821-869.

Pages 6-7 & Table 1

(1) Use the correct notation for Km and Vmax. Also, the authors show kinetic parameters and use multiple units (e.g., mmol/L or mM for Km).

Response: Thank you very much for your careful reading of the manuscript and valuable suggestions. We have corrected this in the revised manuscript.

(2) When comparing the catalytic efficiency of enzymes, kcat/Km (or Vmax/Km) is generally used. The authors present a comparison of catalytic activity from results to conclusion. A clarification of what results are being compared is needed.

Response: Thank you for your comments and suggestions. The catalytic activity is assessed by comparing reaction rates.

Page 7 & Figure 3

In Figure 3A, the authors describe the overall structure, but a simple explanation or labeling within the figure should be added.

Response: Thank you very much for your suggestions, we have made modifications to Figure 3A as follows:

Crystal structures of CsAlaDC and AtSerDC.

(A) Dimer structure of CsAlaDC. The color display of the N-terminal domain, large domain, and C-terminal domains of chain A is shown in light pink, khaki and sky blue, respectively. Chain B is shown in spring green. The PLP molecule is shown as a sphere model. The zinc finger structure at the C-terminus of CsAlaDC is indicated by the red box. The gray spheres represent zinc ions, while the red dotted line depicts the coordination bonds formed by zinc ions with cysteine and histidine.

Figures 3F & 4A

In these figures, the two structures are overlaid and compared, but the colors are very similar to see the differences. The authors should use a different color scheme.

Response: Thank you very much for your suggestions, we have made modifications to the Figure 3F & 4A as follows:

(Figure 3F) - The monomers of CsAlaDC and AtSerDC are superimposed.

CsAlaDC is depicted in spring green, while AtSerDC is shown in plum. The conserved amino acid catalytic ring is indicated by the red box. (Figure 4A) - Superposition of substrate binding pocket amino acid residues in CsAlaDC and AtSerDC. The amino acid residues of CsAlaDC are shown in spring green, the amino acid residues of AtSerDC are shown in plum, with the substrate specificity-related amino acid residue highlighted in a red ellipse.

Pages 7 & 8

Figures 3 and 4 do not include illustrations of what the authors describe in the text. The reader will not be able to understand the descriptions until they download and view the structures themselves. The authors should create additional figures to make it easier for readers to understand the structures.

Response: Thank you very much for your suggestions, we have included supplementary figure 1 in the revised manuscript, which presents more elaborate structural depictions of the two proteins.

Pages 9 & 10

"This result suggested this Tyr is required for the catalytic activity of CsAlaDC and AtSerDC."

The author's results are interesting, but it is recommended to perform the experiments in a specific order. First, experiments should determine whether mutagenesis affects the protein's stability (e.g., CD, as discussed earlier), and second, whether mutagenesis affects ligand binding (e.g., ITC, SPR, etc.), before describing how site-directed mutagenesis alters enzyme activity. In particular, the authors' hypothesis would be much more convincing if they could show that the ligand binding affinity is similar between WT and mutants.

Response: Thank you for your insightful feedback on our manuscript, which we greatly appreciate. Your suggestion to methodically sequence the experiments provides a clear pathway to bolster the strength and conclusiveness of our results.

We agree that it is crucial to first assess the stability of the mutant proteins, as changes therein could inadvertently affect catalytic activity. To this end, we have employed circular dichroism (CD) to study the potential structural alterations in the proteins induced by mutations. The experimental results are shown in the following figure:

(A) Circular Dichroism Spectra of CsAlaDC (WT).

(B) Circular Dichroism Spectra of CsAlaDC (Y336F). (C) Circular Dichroism Spectra of CD of AtSerDC (WT). (D) Circular Dichroism Spectra of AtSerDC (Y341F).

The experimental results indicate that the secondary structure of the mutant proteins remains unchanged, which means the mutations do not alter the protein's stability.

The ligand PLP forms a Schiff base structure with the ε-amino group of a lysine residue in the protein, with maximum absorbance around 420-430 nm. Since we have already added PLP during the protein purification process, as long as the absorbance of mutant proteins and wild-type proteins is the same at 420-430 nm at equivalent concentrations, it indicates that the mutant proteins do not affect the binding of the ligand PLP. Therefore, we scanned the UV-visible absorption spectra of both the wild-type and mutant proteins, and the results are as presented in the following figure:

(A) UV-Visible Absorption Spectra of CsAlaDC (WT) compared to CsAlaDC (Y336F).

(B) UV-Visible Absorption Spectra of AtSerDC (WT) compared to AtSerDC (Y341F).

The mutant protein and the wild-type protein exhibit similar absorbance at 420-430 nm, indicating that the mutation does not affect the binding of PLP to the protein.

The above experiments have confirmed that the mutations do not significantly affect the stability of the protein or the affinity for the ligand, so we can more confidently attribute changes in enzyme activity to the specific role of the tyrosine residue in question. We believe this comprehensive approach will substantiate our hypothesis and illustrate the necessity of this Tyr residue for the catalytic activity of CsAlaDC and AtSerDC enzymes.

Figure 3

In the 3D structure figure provided by the authors, the proposed reaction mechanism of the enzyme and the involved amino acids are not included. Can the authors add a supplementary figure with a schematic drawing that includes more information, such as distances?

Response: Thank you for your valuable feedback on our manuscript. We completely agree that a schematic drawing with additional details, including distances, would enhance the clarity and understanding of the enzymatic mechanism. In response to your suggestion, we have added a supplementary figure 2 in the revised manuscript that accurately illustrates the proposed reaction pathway, highlighting the key amino acids involved.

Page 10

"The results showed that 5 mM L-DTT reduced the relative activity of CsAlaDC and AtSerDC to 22.0% and 35.2%, respectively"

The authors primarily use relative activity to compare WT and mutants. Can the authors specify the exact experiments, units, and experimental conditions? Is it Vmax or catalytic efficiency? If so, under what specific experimental conditions?

Response: Thank you for your attention and review of our research paper, we appreciate your suggestions and feedback. The experimental protocol employed to evaluate the influence of DTT on protein catalytic efficiency is outlined as follows:

The 100 μL reaction mixture, containing 20 mM substrate (Ala or Ser), 100 mM potassium phosphate, 0.1 mM PLP, 5 mM L-DTT, and 0.025 mM purified enzyme, was prepared and incubated at standard conditions (45 °C and pH 8.0 for CsAlaDC for 5 min, 40 °C and pH 8.0 for AtSerDC for 2 min). DTT is absent as a control in the reaction system. Then the reaction was stopped with 20 μL of 10% trichloroacetic acid. The product was derivatized with 6-aminoquinolyl-N-hydroxy-succinimidyl carbamate (AQC) and subjected to analysis by UPLC. All enzymatic assays were performed in triplicate.

However, due to the unknown mechanism of DTT inhibition on protein activity, we have removed this part of the content in the revised manuscript.

Pages 10-12

The identification of 'Phe106 in CsAlaDC' and 'Tyr111 in AtSerDC,' along with the subsequent mutagenesis and enzymatic activity assays, is intriguing. However, the current manuscript lacks an explanation and discussion of the underlying reasons for these results. As previously mentioned, it would be helpful to gain insights and analysis from WT-ligand and mutant-ligand binding studies (e.g., ITC, SPR, etc.). Furthermore, the authors' analysis would be more convincing with accompanying structural analysis, such as steric hindrance analysis.

Response: Thank you for your insightful comments and constructive feedback on our manuscript. We appreciate the interest you have expressed in the identification of 'Phe106 in CsAlaDC' and 'Tyr111 in AtSerDC' and their functional implications based on mutagenesis and enzymatic assays.

In order to investigate the binding status of the mutant protein and the ligand PLP,we scanned the UV-visible absorption spectra of both the wild-type and mutant proteins, and the results are as presented in the following figure:

(A) UV-Visible Absorption Spectra of CsAlaDC (WT) compared to CsAlaDC (F106Y).

(B) UV-Visible Absorption Spectra of AtSerDC (WT) compared to AtSerDC (Y111F).

The mutant protein and the wild-type protein exhibit similar absorbance at 420-430 nm, indicating that the mutation does not affect the binding of PLP to the protein. Therefore, we can conclude that the change in activity of the mutant protein is caused by the substitution of the amino acid at that site, i.e., the amino acid at that site affects substrate specificity. By combining the structure of the two proteins, we can see that the Lys at position 111 of AtSerDC is a hydrophilic amino acid, which increases the hydrophilicity of the active site, and thus the substrate is the hydrophilic amino acid Ser. In contrast, the amino acid at the corresponding site in CsAlaDC is Phe, which, lacking a hydroxyl group compared to Lys, increases the hydrophobicity of the active site, making the substrate lean towards the hydrophobic amino acid Ala. We have added a discussion of the potential reasons for this result to the revised manuscript's discussion section.

Page 5 & Figure 1B

"As expected, CsSerDC was most closed to AtSerDC, which implies that they shared similar functions. However, CsAlaDC is relatively distant from CsSerDC."

In Figure 1B, CsSerDC and AtSerDC are in different clades, and this figure does not show that the two enzymes are closest. To provide another quantitative comparison, please provide a matrix table showing amino acid sequence similarities as a supplemental table.

Response: Many thanks for your constructive suggestion. We added a matrix table showing amino acid sequence similarities in the supplemental materials. The results showed that the similarity of amino acid sequences between CsSerDC and AtSerDC is 86.21%, which is higher than that between CsAlaDC and CsSerDC (84.92%). This data exactly supports the description of Figure 1B. We added the description of the amino acid sequence similarities analysis in the revised manuscript. The description of "As expected, CsSerDC was most closed to AtSerDC, which implies that they shared similar functions. " is not accurate enough, so we revised it to "As expected, CsSerDC was closer to AtSerDC, which implies that they shared similar functions.", in the revised manuscript.

Page 5 & Figure 1C

Figure 1C, which shows a multiple sequence alignment with the amino acid sequences of the 6 SerDCs and CsAlaDC, clearly shows the differences between the sequences of AlaDC and other SerDCs. However, the authors' hypothesis would be more convincing if they showed that this difference is also conserved in AlaDCs from other plants. Can the authors show a new multiple-sequence alignment by adding more amino acid sequences of other AlaDCs?

Response: Thank you for your comments and suggestions. We aim to discover additional alanine decarboxylase. However, at present, the only experimentally confirmed alanine decarboxylase is CsAlaDC. No experimentally verified alanine decarboxylases have been found in other plant species.

Figure 5A

Figure 5A is missing the error bar.

Response: Figure 5A serves as a preliminary screening for these mutants, without conducting repeated experiments. Subsequently, only the L110F and P114A mutants, which exhibited significantly improved activity, underwent further experimental verification to confirm their enhanced functionality.

Version 2

Abstract

Abstract

Ethylamine (EA), the precursor of theanine biosynthesis, is synthesized from alanine decarboxylation by Alanine Decarboxylase (AlaDC) in tea plants. AlaDC evolves from Serine Decarboxylase (SerDC) through neofunctionalization and has lower catalytic activity. However, lacking structure information hinders the understanding of the evolution of substrate specificity and catalytic activity. In this study, we solved the X-ray crystal structures of AlaDC from Camellia sinensis (CsAlaDC) and SerDC from Arabidopsis thaliana (AtSerDC). Tyr341 of AtSerDC or the corresponding Tyr336 of CsAlaDC is essential for their enzymatic activity. Tyr111 of AtSerDC and the corresponding Phe106 of CsAlaDC determine their substrate specificity. Both CsAlaDC and AtSerDC have a distinctive zinc finger that contributes to the activity and has not been identified in any other Group II PLP-dependent amino acid decarboxylases. Based on the structure comparison and evolution analysis AlaDC or SerDC led us to perform mutation screen of CsAlaDC. The results indicated that the mutation of L110F or P114A in the CsAlaDC dimerization interface significantly improved the catalytic activity by 110% and 59%, respectively. Combining double mutant CsAlaDCL110F/P114A with theanine synthetase increased 672% of theanine production in an in vitro system. This study provides the structural basis for the substrate selectivity and catalytic activity of CsAlaDC and AtSerDC and sheds light on a new direction for the efficient industrial synthesis of theanine.

Introduction

Tea is one of the most popular natural non-alcoholic beverages consumed worldwide. About 2 billion cups of tea are consumed worldwide daily (1). The popularity of tea is determined by its favorable flavor and numerous health benefits (2–4). The flavor and health-beneficial effects of tea are conferred by the abundant secondary metabolites, including catechins, caffeine, theanine, volatiles, etc (5). Theanine (γ-glutamylethylamide) is a unique non-protein amino acid and the most abundant free amino acid in tea plants (Camellia sinensis). It accounts for more than 50% of the total free amino acids and approximately 1-2% of the dry weight of the new shoots of tea plants (6). Theanine is the secondary metabolite conferring the umami taste of tea infusion and also balances the astringency and bitterness of tea infusion caused by catechins and caffeine (7). It has also many health-promoting functions, such as reducing stress and anxiety, improving mood and cognition, protecting the nervous system, etc (6, 8–13). Therefore, theanine content is highly correlated with green tea quality (14).

Theanine is synthesized from ethylamine (EA) and glutamate (Glu) by theanine synthetase (15, 16). Importantly, the large amount of theanine biosynthesis is determined by the high availability of EA in tea plants (17). Therefore, the evolution of EA biosynthesis in tea plants provided basis for theanine biosyntheis and the formation of tea quality. EA is synthesized from alanine decarboxylation by alanine decarboxylase (CsAlaDC) (18) (Figure 1A). Indeed, CsAlaDC expression level and catalytic activity largely determine the theanine accumulation level in tea plants (19). As a novel gene, CsAlaDC had not been reported in other plant species before it was identified in tea plants. Previous studies indicated that AlaDC originated from the serine decarboxylase gene (SerDC) by gene duplication in tea plants (20). However, despite sharing highly conserved amino acid sequences, CsAlaDC and CsSerDC specifically catalyze the decarboxylation of alanine and serine, respectively (20) (Figure 1A, 1C). In addition, CsAlaDC exhibited a much lower enzymatic activity compared with CsSerDC (20). However, the structural basis and key amino acids underlying the evolution of the substrate specificity and enzymatic activity of CsAlaDC are unknown.

Metabolic pathways and phylogeny analysis of SerDC and AlaDC in plants.

(A) The decarboxylation of serine and alanine in plants. SerDC, serine decarboxylase; AlaDC, alanine decarboxylase. (B) Phylogenetic analysis of the SerDC homologous proteins in Algal, Pteridohyta, Bryophyta, Dicots and Monocots. The phylogenetic tree was performed using the neighbor-joining method in MAGA 7.0 software and bootstrap values based on 1000 replicates were displayed at branch points. These proteins were represented by their GenBank accession number. (C) Multiple alignment of the amino acid sequences of AlaDC and SerDCs. Primary (100%), secondary (80%), and tertiary (60%) conserved percent of similar amino acid residues were shaded in deep blue, light blue and cheer red, respectively. The multiple alignment was constructed by MAGA 7.0 software by ClustalW method, and GeneDOC software was used to visualize multiple sequence alignments. “*” indicated amino acid residues mutated only in CsAlaDC.

SerDC belongs to the Group II pyridoxal-5-phosphate (PLP)-dependent decarboxylase superfamily (21). These Group II amino acid decarboxylases produce many bioactive secondary metabolites and signaling molecules (22–24). In plants, SerDC catalyzes the biosynthesis of ethanolamine from serine, which was first functionally characterized in Arabidopsis thaliana (25) (Figure 1A). Ethanolamine is an important metabolite in the synthesis of phosphatidylethanolamine and phosphatidylcholine, two main phospholipids involved in maintaining eukaryotic membrane structures in eukaryotic cell membranes (26–28). Furthermore, the study has shown that SerDC is essential in the embryogenesis of Arabidopsis (29). SerDC is of vital importance for plant growth and development, but the mechanism of substrate recognition and catalytic activity of SerDC also remains unclear.

In this work, we attempt to understand the mechanism of functional divergence of plant AlaDC and SerDC by analyzing their structural characteristics. Here, we obtained the X-ray crystal structure of CsAlaDC and AtSerDC. According to the crystal structure, we found a distinctive zinc finger structure located at both CsAlaDC and AtSerDC, which has not been identified in any other Group II PLP-dependent amino acid decarboxylases that have been previously characterized. By comparing the substrate binding pockets, we identified Phe106 of CsAlaDC and Tyr111 of AtSerDC as the crucial sites for substrate specificity. By conducting mutation screening based on the protein structures, we identified the amino acids repressing the catalytic activity and discovered that CsAlaDCL110F/P114A exhibited a 2.3-fold increase in catalytic activity compared to that of CsAlaDC and enhanced the engineering of theanine production in vitro.

Results

Enzymatic properties of CsAlaDC, AtSerDC, and CsSerDC

CsAlaDC originates from CsSerDC by gene duplication and neofunctionalization in tea plants, but they catalyze different metabolic processes (Figure 1A). A phylogenetic tree was constructed to further reveal the evolutionary relationship between CsAlaDC and its homologs in Algal, Pteridohyta, Bryophyta, Monocots and Dicots (Figure 1B). Phylogenetic analysis indicated that CsAlaDC is homologous with SerDCs in Dicots plants. As expected, CsSerDC was closer to AtSerDC, which implies that they shared similar functions. However, CsAlaDC is relatively distant from CsSerDC. Amino acid sequence similarity analysis also showed that CsSerDC and AtSerDC had higher smilarity than that between CsAlaDC and CsSerDC (Supplementary table 1). Next, we performed a multiple sequence alignment of the amino acid sequences of the 6 SerDCs from kiwifriut (Actinidia chinensis), grape (Vitis vinifera), coffee (Coffea eugenioides), cocoa (Theobroma cacao), Arabidopsis and tea plants and CsAlaDC (Figure 1C). The results showed that the amino acid sequences of CsAlaDC and SerDCs were highly conserved, but CsAlaDC has amino acid mutations in highly conserved regions compared with SerDCs.

Next, to verify the substrate specificity and enzyme activity, we conducted enzyme activity detection and enzyme kinetics analysis for CsAlaDC, CsSerDC and AtSerDC. The 5′ truncated CsAlaDC and AtSerDC were constructed into the pET22b and pET28a expression vectors to generate recombinant plasmids pET22b-CsAlaDC and pET28a-AtSerDC, correspondingly. Additionally, the full-length protein of CsSerDC was inserted into the pET28a expression vector to generate recombinant plasmid pET28a-CsSerDC. Subsequently, the recombinant proteins were expressed in Escherichia coli and purified via nickel affinity chromatography. The purified proteins were examined through SDS-PAGE analysis (Figure 2A). To investigate the substrate specificity of CsAlaDC, AtSerDC, and CsSerDC, enzyme activity assays were conducted using Ala and Ser as substrates, followed by product identification via UPLC analysis. The results showed that CsAlaDC can effectively catalyzed the decarboxylation of alanine to generate EA, whereas its catalytic efficacy towards serine was relatively inferior. Conversely, AtSerDC and CsSerDC selectively catalyzed serine decarboxylation to yield ethanolamine but did not exhibit activity towards alanine (Figure 2B). These findings confirmed that CsAlaDC and SerDCs do not have promiscuous decarboxylase activity on Ala and Ser.

Purification and characterization of CsAlaDC, AtSerDC, and CsSerDC.

(A) Identification of CsAlaDC, AtSerDC, and CsSerDC using SDS-PAGE. (B) Detection of enzyme activities of CsAlaDC, AtSerDC, and CsSerDC by UPLC. (C) Reaction rates of substrates with different concentrations catalyzed by CsAlaDC, AtSerDC, and CsSerDC.

The kinetic properties of CsAlaDC, AtSerDC, and CsSerDC were determined through the use of corresponding substrates (Figure 2C), and the results are presented in Table 1. The kinetic parameter Km of CsAlaDC was determined to be 1.215 mM, which is similar to that of AtSerDC and CsSerDC at 1.522 mM and 2.364 mM, respectively. However, AtSerDC has a Vmax of 7.053 μmol/L/s, which is 4-fold that of CsAlaDC’s Vmax of 1.709 μmol/L/s; while CsSerDC has a Vmax of 9.031 μmol/L/s, which is 5.3-fold that of CsAlaDC. These findings indicated that the catalytic efficiency of CsAlaDC is considerably lower than that of both CsSerDC and AtSerDC. The observed disparity in Vmax suggests that CsAlaDC and SerDCs may exhibit discrepancies in substrate-to-product conversion rates under saturated substrate conditions. It is plausible that SerDCs possess a higher turnover rate, facilitating the rapid conversion of substrate to product and subsequent release of the enzyme. Another possibility is that SerDCs demonstrate enhanced stability, thereby maintaining superior catalytic activity even at higher substrate concentrations, consequently leading to a higher Vmax value. Due to the similarity in catalytic activity between CsSerDC and AtSerDC, we chose the more representative AtSerDC for further analysis.

The overall structures of CsAlaDC and AtSerDC

To enhance our comprehension of CsAlaDC and AtSerDC, we conducted structural analyses of these two proteins. Through optimization of crystallization conditions, we successfully determined the crystal structures of CsAlaDC, CsAlaDC-EA, and AtSerDC at resolutions of 2.50 Å, 2.60 Å and 2.85 Å, respectively (Supplementary table 2).

The overall structure of CsAlaDC and AtSerDC is homodimeric with two subunits exhibiting an asymmetrical arrangement (Figure 3A and 3D). The monomer of CsAlaDC or AtSerDC is divided into three distinct structural domains: an N-terminal domain (N-terminal −104 aa in CsAlaDC, N-terminal −109 aa in AtSerDC), a large domain (105-364 aa in CsAlaDC, 110-369 aa in AtSerDC), and a small C-terminal domain (365 aa-C-terminus in CsAlaDC, 370-C-terminus as in AtSerDC), colored light pink, khaki, and sky blue, respectively. Compared to the full-length protein, the N-terminal structures of CsAlaDC and AtSerDC were truncated by 60 and 65 amino acid residues, respectively. The N-terminal domain of the truncated protein contains a long α-helix, which is connected to the large domain through a long loop. The large domain contains a seven-stranded mixed β-sheet surrounded by eight α-helices, while the C-terminal domain is composed of a three-stranded antiparallel β-sheet and three α-helices (Supplementary figure 1). The catalytic site of the enzyme is positioned within a superficial crevice at the junction of two subunits forming a dimeric arrangement. The amino acid residues derived from both subunits participate in the binding of the PLP, effectively anchoring it to the active site. In the catalytic site, one monomer accommodates PLP, while the other monomer also contributes to PLP binding and enzymatic function (Figure 3A and 3D).

Crystal structures of CsAlaDC and AtSerDC.

(A) Dimer structure of CsAlaDC. The color display of the N-terminal domain, large domain, and C-terminal domains of chain A is shown in light pink, khaki and sky blue, respectively. Chain B is shown in spring green. The PLP molecule is shown as a sphere model. The zinc finger structure at the C-terminus of CsAlaDC is indicated by the red box. The gray spheres represent zinc ions, while the red dotted line depicts the coordination bonds formed by zinc ions with cysteine and histidine. (B) The 2Fo-Fc electron density maps of K309-PLP-EA (contoured at 1σ level). The PLP is shown in violet, the K309 is shown in spring green, and the EA is shown in lightblue. (C) Active center of the CsAlaDC-EA complex, with hydrogen bonds denoted by black dotted lines. “*” denotes the amino acids on adjacent subunits. (D) Dimer structure of AtSerDC. The color display of the N-terminal domain, large domain, and C-terminal domains of chain A is shown in light pink, khaki and sky blue, respectively. Chain B is shown in cyan. The PLP molecule is shown as a sphere model. The zinc finger structure at the C-terminus of AtSerDC is indicated by the red box. The gray spheres represent zinc ions, while the red dotted line depicts the coordination bonds formed by zinc ions with cysteine and histidine. (E) Active center of the AtSerDC, with hydrogen bonds denoted by black dotted lines. “*” denotes the amino acids on adjacent subunits. (F) The monomers of CsAlaDC and AtSerDC are superimposed. CsAlaDC is depicted in spring green, while AtSerDC is shown in plum. The conserved amino acid catalytic ring is indicated by the red box. (G) Amino acid residues of the active center in CsAlaDC apo and CsAlaDC-EA complex are superimposed. CsAlaDC apo is shown in floral white, while CsAlaDC-EA complex is shown in spring green. (H) The relative activity of wild-type CsAlaDC and its Y336F mutant (left), as well as wild-type AtSerDC and its Y341F mutant (right) is shown.

Notably, our investigation has revealed the presence of a distinctive zinc finger structure (as depicted in Figure 3A and 3D) located at the C-terminus of CsAlaDC and AtSerDC. This structure is composed of a loop structure spanning 17 amino acid residues, wherein coordination of Zn2+ is facilitated by three Cys residues and one His residue. Importantly, this particular configuration is exclusive to the two proteins under examination and has not been identified in any other Group II PLP-dependent amino acid decarboxylases that have been previously characterized (30, 31).

The crystal structures of CsAlaDC-EA complex and AtSerDC were obtained and further analyzed. In the former complex, PLP bound to Lys309 via a Schiff base linkage (internal aldimine form), and the pyridine moiety of PLP is positioned between the imidazole ring of His196 and the methyl group of Ala279 in a parallel orientation to the imidazole ring. The carboxylic group of Asp277 stabilizes the N1 of PLP by a salt bridge interaction, providing the latter with the strong electrophilicity necessary to stabilize carbon ion intermediates during enzymatic catalysis. The O3 forms hydrogen bonds with Thr247 and Lys309. Additionally, His308, Gly169, Thr170, Lys309, and Ser347*(“*” denotes the amino acids on adjacent subunits) establish a hydrogen bonding network with the phosphate group of PLP, consequently fortifying its attachment to the protein. (Figure 3C). Similarly, in the latter structure, the pyridine ring of PLP was sandwiched between the imidazole ring of His201 and the methyl group of Ala284. The N1 of PLP formed a salt bridge with Asp282, while the O3 formed hydrogen bonds with Thr252. The phosphate group of PLP established hydrogen bonding interactions with Lys314, His313, Gly174, Thr175, and Ser352* (Figure 3E). The aforementioned amino acid residues are observed across various Group II PLP-dependent amino acid decarboxylases and exhibit considerable stability (see further discussion below), suggesting that they could share common features in terms of catalytic mechanisms.

Factors affecting CsAlaDC and AtSerDC activity

The Group II PLP-dependent amino acid decarboxylases are dimeric enzymes with each monomer containing an active site located within a shallow cavity (32). A flexible loop of amino acids originating from one monomer extends into the active site of the other monomer and plays a crucial role in catalysis. The catalytic loop in CsAlaDC, consisting of amino acid residues 328-341, is well-structured and exhibits clear electron density. However, the corresponding loop in AtSerDC is disordered (Figure 3F). The aforementioned loop harbors a conserved Tyr residue that is believed to function as a proton donor of the carbanion in the catalytic process (33). In particular, CsAlaDC Tyr336 and AtSerDC Tyr341 correspond to that Tyr residue (Supplementary figure 2). Upon superimposing the active site residues of CsAlaDC apo and CsAlaDC-EA complex, it was observed that the hydroxyl group of Tyr336* in the CsAlaDC-EA complex exhibits a substantial 60-degree deviation along the Cα-Cβ bond (Figure 3G). To assess the function of the Tyr, we substituted the corresponding Tyr residues in CsAlaDC and AtSerDC with Phe. The resulting CsAlaDCY336F and AtSerDCY341F mutants were then exposed to L-alanine and L-serine, respectively. We measured the production of EA or ethanolamine in the reaction mixtures. Our findings indicate that these mutants catalyze abortive decarboxylation, as opposed to wild-type CsAlaDC and AtSerDC, as evidenced by the absence of detectable EA and only a small amount of ethanolamine observed in the reaction mixture (Figure 3H). This result suggested this Tyr is required for the catalytic activity of CsAlaDC and AtSerDC.

Identification of key amino acids for the substrate specificity

Through the superposition of the amino acid residues present in the substrate binding pocket of CsAlaDC and AtSerDC, it was observed that the residues were identical, except for Phe at position 106 in CsAlaDC (Figure 4A). In AtSerDC, this position is occupied by Tyr at 111, implying that this specific amino acid residue is related to substrate specificity.

Key amino acid residues for substrate recognition.

(A) Superposition of substrate binding pocket amino acid residues in CsAlaDC and AtSerDC. The amino acid residues of CsAlaDC are shown in spring green, the amino acid residues of AtSerDC are shown in plum, with the substrate specificity-related amino acid residue highlighted in a red ellipse. (B) Active-site-lining residues from Embryophyta AlaDC were identified. The height of the residue label displays the relative amino acid frequency. (C) Histogram showing the distribution of the number of key motifs. (D) Histogram showing the number of key motifs in different plant orders. (E) Relative enzyme activities of wild-type CsAlaDC and mutant protein CsAlaDCF106Y against Ala substrate (columns 1 and 2), and enzyme activities of wild-type AtSerDC and various AtSerDC mutant proteins against Ala substrate (columns 3-9) are presented. The percentage graph shows the relative activity of each protein compared to wild-type CsAlaDC activity (taken as a 100% benchmark). (F) Relative enzyme activities of wild-type AtSerDC and AtSerDC mutant proteins (columns 1-7) against Ser substrates, and enzyme activities of wild-type CsAlaDC and mutant protein CsAlaDCF106Y against Ser substrates(columns 8, 9) were measured. The percentage graph shows the relative activity of each protein compared to the wild-type AtSerDC (taken as a 100% benchmark). Three independent experiments were conducted. (G) The EA contents of AtSerDC and its mutant AtSerDCY111F in N. benthamiana. (H) The EA contents of CsAlaDC and its mutant CsAlaDCF106Y in N. benthamiana. The significance of the difference (P<0.05) was labeled with different letters according to Duncan’s multiple range test.

To obtain more insights into the role of specific amino acid residue in substrate specificity, we utilized amino acid sequences of AtSerDC to identify all potential serine decarboxylases in Embryophyta and constructed phylogenetic trees (Supplementary figure 3). We further identified conserved motifs in these SerDC homologs in Embryophyta (Figure 4B). Importantly, the Phe106 of CsAlaDC or Tyr111 of AtSerDC locates in a conserved motif (Figure 4B). In this conserved motif, the first two amino acids Y and P are 100% conserved, but the third amino acid is variable and can be Y, T, F, V, A, I, or other amino acids (Figure 4B and C); interestingly, the Phe106 of CsAlaDC and Tyr111 of AtSerDC locates at this variable amino acid position. For these first three amino acids, the YPY was identified in 83.7% of the SerDC homologs, while the YPF motif was found in 2.3% of these SerDC homologs (Figure 4C). These analyses provided more insights into the role of Phe106 of CsAlaDC and Tyr111 of AtSerDC in the substrate specificity.

To verify the role of Phe106 of CsAlaDC or Tyr111 of AtSerDC in substrate specificity, we mutated Phe106 of CsAlaDC into Tyr (CsAlaDCF106Y) and Tyr111 of AtSerDC into Phe (AtSerDCY111F) and tested enzymatic activity in vitro. The results showed that CsAlaDCF106Y totally lost the alanine decarboxylase activity, and AtSerDCY111F gained alanine decarboxylase activity (Figure 4E). We also performed other mutations of Tyr111of AtSerDC, including AtSerDCY111A, AtSerDCY111I, AtSerDCY111L, AtSerDCY111V and AtSerDCY111W. Within these mutations, only AtSerDCY111I exhibited a low level of alanine decarboxylase activity (Figure 4E). These results verified that the Phe106 of CsAlaDC is required for the alanine decarboxylase activity from serine decarboxylase activity.

On the other side, both CsAlaDC and CsAlaDCF106Y exhibit a very low level of serine decarboxylase activity (Figure 4F). Moreover, compared with AtSerDC, the serine decarboxylase activity of AtSerDCY111F was about 30% of AtSerDC; AtSerDCY111I retained lower than 5% of serine decarboxylase activity of AtSerDC; while other mutations, including AtSerDCY111A, AtSerDCY111L, AtSerDCY111V and AtSerDCY111W abolished the serine decarboxylase activity (Figure 4F). These results suggested the Tyr111 of AtSerDC is also important for the substrate specificity of AtSerDC.

To further verify that Phe106 of CsAlaDC and Tyr111 of AtSerDC were key amino acid residues determining its substrate recognition in planta, Nicotiana benthamiana transient expression system was carried out. To this end, A. tumefaciens strain GV3101 (pSoup-p19), carrying recombinant plasmid, was infiltrated into leaves of 5-week-old N. benthamiana plants, and the pCAMBIA1305 empty vector was used as the control (EV). Relative mRNA levels were detected (Supplementary figure 4), indicating that they had been successfully overexpressed in tobacco. Here, we found that a high level of EA was produced in the CsAlaDC-expressing tobacco leaves; while no EA product was detected in tobacco leaves infiltrated with the mutant CsAlaDCF106Y. In addition, we did not detect EA products in the AtSerDC-expressing tobacco leaves, while a high level of EA was detected in tobacco leaves infiltrated with mutant AtSerDCY111F. As anticipated, EA was not detected in tobacco leaves infiltrated with EV (Figure 4G and H). These results further verified the critical role of Phe106 in the substrate specificity of CsAlaDC.

Key amino acids for the evolution of CsAlaDC enzymatic activity

CsAlaDC and AtSerDC have a high sequence similarity of 74.5% and a nearly identical structure with an RMSD (root mean square deviation) of only 0.77Å for monomer structures. However, the two enzymes catalyze different amino acid decarboxylation reactions, and AtSerDC exhibits significantly higher activity than CsAlaDC (Figure 2B and C). To elucidate this observation, we proposed a prediction test: EA, generated via Ala decarboxylation by CsAlaDC, can be toxic and harmful to plants if accumulated excessively. Thus, during the evolution of plant serine decarboxylase into alanine decarboxylase, the enzyme has evolved not just to alter the substrate preference but also to reduce catalytic activity to control EA production within a suitable range. Based on this hypothesis, we suggest that mutating specific amino acids in CsAlaDC to those corresponding amino acids in SerDCs could enhance its activity, and the results could provide insights into the evolution of CsAlaDC enzymatic activity.

Dimerization is essential to AlaDC/SerDC activities because the active site is composed of residues from two monomers. To identify the amino acids repressing the enzymatic activity of CsAlaDC during evolution from SerDC, we analyzed the crystal structures and the amino acids at the dimer interface between CsAlaDC and AtSerDC. This analysis revealed that the amino acid residues at positions 66, 97, 110, 114, 116, 117, 122, 315 and 345 are different in CsAlaDC and SerDCs (Figure 1C). Therefore, we mutated these amino acids of CsAlaDC into those in the corresponding positions of AtSerDC and CsSerDC, and performed enzyme activity assays (Figure 5A). The results demonstrated that CsAlaDCL110F and CsAlaDCP114A exhibited significantly enhanced enzyme activity compared to the wild-type CsAlaDC, with a 2.1-fold and 1.59-fold increase, respectively. Furthermore, the catalytic activity of the CsAlaDCL110F/P114A double mutant exhibits a remarkable 2.3-fold increase compared to that of the wild-type protein (Figure 5B). These findings suggested a critical role of Leu110 and Pro114 of CsAlaDC in the evolution of enzymatic activity and provided a basis to improve CsAlaDC activity. It is possible that these amino acid residues could potentially augment the hydrophobic nature of the protein dimer interface.

Mutations enhance CsAlaDC enzyme activity and theanine synthesis in vitro.

(A) Relative enzyme activities of CsAlaDC mutant proteins against Ala substrate. (B) Relative enzyme activities of CsAlaDCL110F, CsAlaDCP114A, and CsAlaDCL110F/P114A against Ala substrate. (C) Histogram showing the relative content of theanine resulting from different combinations of alanine decarboxylase and theanine synthetase. Three independent experiments were conducted.

In Vitro synthesis of theanine

The biosynthetic pathway of theanine in tea plants comprises two sequential enzymatic steps: alanine decarboxylase mediates alanine decarboxylation to yield EA, which is subsequently utilized by theanine synthetase to effectuate the condensation reaction with glutamate to synthesize theanine. By conducting mutation screening, we discovered a CsAlaDC mutant protein (L110F/P114A) that exhibited a 2.3-fold higher catalytic activity compared to the wild-type protein. Subsequently, we employed an in vitro theanine synthesis system utilizing CsAlaDC and either glutamine synthetase (PtGS [Pseudomonas syringae pv. syringae]) or gamma-glutamate methylamine ligase (GMAS [Methylovorus mays]) which have the activity to synthesize theanine from EA and Glu (34, 35).

The results illustrated the successful synthesis of theanine using CsAlaDC in conjunction with two theanine synthetases, employing Ala and glutamate as substrates (Figure 5C). The theanine content generated by the combination of CsAlaDCL110F and GMAS, as well as CsAlaDCL110F/P114A and GMAS, was 4.57-fold and 6.72-fold higher than the content produced in the combination of wild-type CsAlaDC and GMAS, respectively. Similarly, when combined with the PsGS, comparable outcomes are observed as well. The theanine content resulting from the combination of CsAlaDCL110F and PsGS, as well as CsAlaDCL110F/P114A and PsGS, exhibit enhancements of 1.62-fold and 4.33-fold compared to the wild-type protein combination, respectively (Figure 5C). Hence, the utilization of CsAlaDCL110F/P114A could effectively enhance the theanine production yield and thus, holds potential for large-scale engineering production of theanine.

Discussion

Comparison of CsAlaDC and AtSerDC with other decarboxylases

The Dali search (36) revealed that AtSerDC and CsAlaDC exhibit structural similarity to several other decarboxylases, including 7CIG (MetDC [Streptomyces sp. 590]), 7ERV (HisDC1 [Photobacterium phosphoreum]), 6KHO (TrpDC [Oryza sativa Japonica Group]), 4E1O (HisDC2 [Homo sapiens]), 6JY1 (AspDC [Methanocaldococcus jannaschii]), and 5GP4 (GluDC [Levilactobacillus brevis]). The monomers of these amino acid decarboxylases all consist of three characteristic domains (Figure 3 and Supplementary figure 6). The C-terminal domain, also known as the small domain, is composed of three antiparallel β-sheets and three α-helices, facing toward the opposite monomers. The large domain contains seven stranded β-sheets surrounded by eight α-helices, exhibiting characteristics typical of Type I enzymes. This domain includes a conserved PLP-binding active center and amino acid residues that are conserved across species. The N-terminal domain consists of elongated α-helices and is connected to the large domain via a loop. The helix of one subunit is antiparallel to the corresponding helix in the neighboring subunit, forming a clamp. Therefore, the N-terminal domain may not be an independent folding unit and is likely only stable in dimers.

The monomeric structure of CsAlaDC and AtSerDC exhibits numerous structural features common to other Group II PLP-dependent amino acid decarboxylases, and the amino acid residues that bind to PLP cofactors are conserved across multiple enzymes (Supplementary figure 5 and 6). Out of the eight proteins analyzed, a total of thirteen amino acid residues were found to be conserved across all sequences (Glu143, Glu171, Lys201, Gly245, Thr247, Asp253, His275, Asp277, Ala279, Pro286, Ser302, Lys309, and Tyr336 in CsAlaDC). Some of them are situated within the active center and play a role in stabilizing PLP or facilitating catalytic reactions. Conversely, other residues reside outside the active center and their function remains unclear. (Supplementary figure 5). PLP is bound to conserved lysine residues in the active center through the Schiff base. The carboxylic group of conserved Asp residues stabilizes protonated pyridine nitrogen of PLP by electrostatic action, providing the latter with strong electrophilicity to stabilize carbanion intermediates during enzymatic catalysis. Another conserved His residue is situated laterally to the PLP, while on the opposite side, there is a conserved Ala residue. The oxygen atom at position 3 of PLP is stabilized by hydrogen bonding with the Thr residue. The phosphate moiety of PLP is additionally secured to the protein through an elongated system of hydrogen bonding. The α5 helix located within the large domain exhibits evolutionary conservation in Group II enzymes, facilitating interaction with the PLP phosphate moiety and conferring appropriate stabilization thereof.

The superimposition of the crystal structures of CsAlaDC and AtSerDC with those of related enzymes suggests that residue Phe106 in CsAlaDC (corresponding to Tyr in AtSerDC and Glu in MetDC) may play a role in determining substrate specificity. This hypothesis was tested via mutagenesis experiments, which revealed that substitution of Phe106 for Tyr rendered CsAlaDC inactive, while substitution of Tyr111 for Phe in AtSerDC enabled alanine decarboxylase activity. Upon contemplation of these experimental outcomes and informed by the elucidated crystal structures of both enzymes, we speculate the hydrophilic nature of the Lys residue in AtSerDC predisposes the active site to preferentially accommodate the hydrophilic amino acid Ser as its substrate. In contrast, the analogous position in CsAlaDC is occupied by Phe, an amino acid devoid of the hydroxyl moiety present in Lys, which augments the hydrophobic character of the substrate-binding site. This structural nuance portends a predisposition for CsAlaDC to select the comparatively hydrophobic amino acid alanine as its suitable substrate. Likewise, the substitution of Glu64 for Ala in MetDC reduced enzyme activity by approximately 50-fold compared to the wild-type (37), bolstering the conclusion that the amino acid residue at this site influences the substrate specificity of the enzyme.

Group II amino acid decarboxylase is characterized by the existence of highly flexible catalytic rings, which are of great significance for the catalytic mechanism of decarboxylase (33). The catalytic rings are located at the dimer interface and extend to the active sites of other monomers in a closed conformation. In the structure of CsAlaDC, the catalytic ring is orderly and allocated, while in the structure of AtSerDC, it is disordered and lacks electron density (Figure 3F). Prior research has established that the conserved amino acid residue Tyr within the ring plays a crucial role in catalysis by donating protons to carbanions of quinone intermediates that arise following decarboxylation (38, 39). Our experiments demonstrate that the substitution of the corresponding Tyr with Phe in the loop renders the protein inactive. Furthermore, alterations in the structure of CsAlaDC when combined with EA provide further evidence supporting this conclusion. Specifically, the structure of CsAlaDC-EA complex displayed a significant conformational change in Tyr336* located in the catalytic site (as depicted in Figure 3G). Comparing the structure of CsAlaDC apo, the CsAlaDC-EA complex exhibited a notable 60-degree shift in the p-hydroxyl group of Tyr336* along the Cα-Cβ bond.

The monomeric configuration of CsAlaDC and AtSerDC is akin to that of other Group II amino acid decarboxylases, except for a conspicuous dissimilarity. Unlike other Group II amino acid decarboxylases, the C-terminal of CsAlaDC and AtSerDC harbors an evident zinc finger structure. We propose that this structure could potentially influence enzyme stability. To test this hypothesis, we truncated the zinc finger structure from both proteins and expressed them. Our findings indicate that, following the excision of the zinc finger structure, CsAlaDC became insoluble, while AtSerDC displayed a similar trait to a certain extent.

Evolution of AlaDC and SerDC

The sequence identity between CsAlaDC and AtSerDC is 74.5%, indicating a significant degree of homology. Moreover, the two proteins exhibit a high level of monomer structural similarity. Notably, they differ in substrate specificity. Using the crystal structure of CsAlaDC and AtSerDC as a basis, we identified pivotal amino acid residues that govern the substrate specificity of serine or alanine decarboxylase. We observed that sequences proximal to these sites exhibit a high degree of conservation in both proteins and possess characteristic motifs (Figure 4B). Moreover, we analyzed serine decarboxylase-like proteins in Embryophyta that are highly homologous to CsAlaDC and constructed phylogenetic trees to gain insight into their evolutionary relationships. Based on our experimental findings and evolutionary evidence, we have identified a conserved YPX motif at the substrate binding pocket. For most plants, the residue X in this motif is predominantly Tyr, which corresponds to the serine decarboxylase protein. However, other residues such as Thr, Phe, Val, Ala, and Ile can also occupy this position (Figure 4B and C). Interestingly, serine decarboxylase-like proteins containing YPF motifs, which have the potential to function as alanine decarboxylases, were found to be distributed throughout the phylogenetic tree, including Asteroideae, Ericales, Chenopodiaceae, Poaceae, and Ranunculales (Supplementary figure 3B). However, these proteins were absent in some more recent species. This observation implies that the emergence of alanine decarboxylase in these particular species could be attributed to convergent evolution.

Moreover, we have observed the enrichment of serine decarboxylase-like proteins that possess a YPT motif within the Fabales. Prior research has demonstrated that XP_004496485-1, which possesses a YPT motif, displays catalytic efficacy toward the processes of decarboxylation and oxidative deamination of Phe, Met, Leu, and Trp (40). Given that the YPT motif is highly conserved and widely distributed in Fabales, serine decarboxylase-like proteins bearing the YPT motif may have developed a unique substrate specificity in Fabales, beyond their conventional decarboxylation functions, as exemplified by XP_004496485-1 protein mentioned above. We speculate that these proteins may be capable of catalyzing other reactions, potentially involving non-protein amino acids or other substances as substrates.

Applied to improve the synthesis of theanine

Theanine is an important indicator of green tea quality. Therefore, improving the synthesis of theanine in tea plants is the focus of research. In this study, through crystal structure analysis and mutation verification, we found that the catalytic activity of CsAlaDCL110F/P114A is 2.3 times higher than that of the wild-type protein, resulting in a more abundant synthesis of theanine in vitro. This gives us great inspiration to improve the synthesis of EA in tea plants by gene editing, thus increasing the content of theanine in tea plants (Figure 5).

Theanine is highly demanded, by the market, due to its health effects and medicinal value, and as a constituent in food, cosmetics and in other fields. To meet the market demand, a variety of methods have been used to acquire theanine, with the main methods including direct extraction, chemical synthesis, biotransformation (microbial fermentation) and plant cell culture. Based on this study, we can use mutation technology to modify engineering bacteria and improve the possibility of synthesis of theanine in bacteria, yeast, model plants, etc, promoting its use for industrial theanine production.

Conclusions

In conclusion, our structural and functional analyses have significantly advanced understanding of the substrate-specific activities of alanine and serine decarboxylases, typified by CsAlaDC and AtSerDC. Critical amino acid residues responsible for substrate selection were identified—Tyr111 in AtSerDC and Phe106 in CsAlaDC— highlighting pivotal roles in enzyme specificity. The engineered CsAlaDC mutant (L110F/P114A) not only displayed enhanced catalytic efficiency but also substantially improved L-theanine yield in a synthetic biosynthesis setup with PsGS or GMAS. Our research expanded the repertoire of potential alanine decarboxylases through the discovery of 13 homologous enzyme candidates across embryophytic species and uncovered a special motif present in serine protease-like proteins within Fabale, suggesting a potential divergence in substrate specificity and catalytic functions. These insights lay the groundwork for the development of industrial biocatalytic processes, promising to elevate the production of L-theanine and supporting innovation within the tea industry.

Materials and Methods

Plants materials

Tobacco (Nicotiana benthamiana) plants were grown in a controlled chamber, under a 16-h light and 8-h dark photoperiod at 25 ℃. Leaves of 5-week-old tobacco plants were used for transient transformation, mediated by Agrobacterium tumefaciens strain GV3101.

Gene cloning and protein expression

The coding sequences of AtSerDC and CsAlaDC, amino acid residues 66-482 and 61-478, respectively, were amplified using PCR and then ligated into the Nde I and Xho I restriction sites of the pET-28a and pET-22b vector, respectively, providing the recombinant vector pET-28a-AtSerDC and pET-22b-CsAlaDC. The cDNAs encoding CsSerDC were amplified using PCR and then ligated into the Nde I and Xho I restriction sites of the pET-28a vector. Gene-specific primers were listed in Supplementary table 3. The recombinant plasmid was transformed into E. coli BL21 (DE3) competent cells. Positive transformants were grown in a 5 mL LB medium containing 30 μg/mL kanamycin or 50 μg/mL ampicillin at 37 °C overnight and then subcultured into an 800 mL LB medium containing the corresponding antibiotic. Protein expression was induced by the addition of 0.2 mM isopropyl-β-d-thiogalactoside (IPTG) for 20 h at 16 ℃ when the optical density (OD) at 600 nm reached 0.6-0.8, harvested by centrifugation at 4 °C and 4,000 rpm for 30 min.

Protein production and crystallization

The cell pellet was suspended in 30 mL of lysis buffer (20 mM HEPES, pH 7.5, 200 mM NaCl, 0.1 mM PLP), disrupted by High-Pressure Homogenizer, and then centrifuged at 16,000 rpm for 30 min at 4 °C to remove the cell debris. The supernatants were purified with a Ni-Agarose resin column followed by size-exclusion chromatography. Before crystallization, purified proteins were concentrated at 10 mg/mL. Crystallization conditions were screened by the sitting-drop vapor diffusion method using the reservoir solutions supplied in commercially available screening kits (Crystal Screen, Crystal Screen 2, PEGRx 1, 2, and SaltRx 1, 2). A droplet made by mixing 1.0 μL of purified AtSerDC or CsAlaDC (10 mg/mL) with an equal volume of a reservoir solution was equilibrated against 100 μL of the reservoir solution at 16℃. The crystal of AtSerDC was obtained using buffer pH 8.0 containing 20% (w/v) PEG400 as a precipitate and 0.2 M CaCl2. The crystal of the CsAlaDC-EA complex was obtained at pH 7.5 containing 2.6 M sodium acetate. The crystal of CsAlaDC was obtained at pH 6.0 containing 3.5 M sodium formate.

Data collection and processing

Crystals were grown by sitting-drop vapor diffusion method at 16 °C. The volume of the reservoir solution was 100 µL and the drop volume was 2 µL, containing 1 µL of protein sample and 1 µL of reservoir solution. The reservoir solution of AtSerDC contained 0.1 M Tris-HCl (pH 8.0), 0.2 M CaCl2, and 20% PEG 400. The reservoir solution of CsAlaDC contained 0.1 M HEPES (pH 7.5), 2.6 M sodium acetate or 0.1 M Bis-Tris (pH 6.0), and 3.5 M sodium formate. Crystals grew in 3-5 days using a protein concentration of 10 mg/mL. Diffraction data were collected at Shanghai Synchrotron Radiation Facility (China). The collected data sets were indexed, integrated, and scaled using the HKL3000 software package. The structure of AtSerDC was solved by molecular replacement using the structure of HisDC (PDB code: 7ERV [Photobacterium phosphoreum]) as the model, and utilizing AtSerDC as a molecular displacement model for both CsAlaDC and the CsAlaDC-EA complex. The AtSerDC, CsAlaDC, and CsAlaDC-EA complex exhibit resolutions of 2.85 Å, 2.50 Å, and 2.60 Å, respectively. The statistics for data collection and processing are summarized in Supplementary table 2.

Enzyme activity assays

Decarboxylase activity was measured by detecting products (EA or ethanolamine) in Waters Acquit ultraperformance liquid chromatography (UPLC) system (20, 41). The 100 μL reaction mixture, containing 20 mM substrate (Ala or Ser), 100 mM potassium phosphate, 0.1 mM PLP, and 0.025 mM purified enzyme, was prepared and incubated at standard conditions (45 °C and pH 8.0 for CsAlaDC, 40 °C and pH 8.0 for AtSerDC for 10 min). Then, the reaction was stopped with 20 μL of 10% trichloroacetic acid. The product was derivatized with 6-aminoquinolyl-N-hydroxy-succinimidyl carbamate (AQC) and subjected to analysis by UPLC. All enzymatic assays were performed in triplicate.

The detection methodology for theanine production remains consistent with the aforementioned approach. The 100 μL reaction mixture, containing 20 mM Ala, 45 mM Glu, 50 mM HEPES (pH 7.5), 0.1 mM PLP, 30 mM MgCl2, 10 mM ATP, 0.03 mM PsGS/GMAS, and 0.025 mM CsAlaDC/CsAlaDC L110F/CsAlaDC L110F/P114A, was prepared and incubated at standard conditions for 1 h. Subsequently, the reaction was terminated via immersion of the reaction vessel in a metal bath at 96 °C for 3 min (35). Theanine was derivatized with AQC and subjected to analysis by UPLC. All enzymatic assays were performed in triplicate.

Site-directed mutagenesis

Site-directed mutagenesis experiment was conducted using a PCR method from the wild-type construct pET-28a-AtSerDC and pET-22b-CsAlaDC, respectively. Dpn I endonucleases were used to digest the parental DNA template. The reaction mixture was used to transform E. coli DH5α competent cells and the plasmids from positive strains were extracted to E. coli BL21 (DE3) for protein expression, purification, and analyzing the enzymatic activity.

In vivo enzyme activity assay in N. benthamiana

The amplified PCR products were fused to the plant expression vector, pCAMBIA1305. Linearization was conducted by restriction digest with Spe I and BamH I. The recombinant colonies were selected for PCR validation on the appropriate antibiotics plate. After validation, the plasmids were electroporated into Agrobacterium tumefaciens strain GV3101(pSoup-p19). Empty pCAMBIA1305 vector, with an intron containing the GFP gene, was treated as the control.

Agrobacterium transient expression assays were performed on 5-week-old N. benthamiana plants. Agrobacterium tumefaciens strain GV3101 (pSoup-p19), carrying the above-described vectors, were cultured in Luria Bertani (LB) medium, containing appropriate antibiotics, at 28 °C. When the absorbance of bacteria colonies reached OD600 = 0.6-0.8, bacterial cells were collected and resuspended in MMA solution (10 mM MgCl2, 10 mM 2-(N-morpholino) ethane sulfonic acid (MES), pH 5.6). After the OD600 of the resuspended bacterial solution was adjusted to approx. 1.0, acetosyringone (AS) was added with a final concentration of 200 µM, and this solution was then incubated, at room temperature, for at least 3 h in darkness. Next, cell suspensions were infiltrated into N. benthamiana leaves with a needle-free syringe. These N. benthamiana leaves were then collected 3 days post infiltration, frozen in liquid nitrogen and stored at −80 °C. Internal EA in N. benthamiana leaves was extracted as previously described (22), and then the solvent was subjected to gas chromatography-mass spectrometry GS-MS system.

Transcript level analysis in N. benthamiana

Total RNA was isolated from samples using the RNAprep Pure Plant Kit (Tiangen, Beijing, China), according to the manufacturer’s protocol. The cDNAs were synthesized using TransScript One-Step gDNA Removal and cDNA Synthesis SuperMix Kit (TransGen Biotech, Beijing, China). The qRT-PCR assays were performed, as previously described (42). Primers used for qRT-PCR assays were listed in Supplementary table 4 and the qRT-PCR was run on a Bio-Rad CFX96TM RT PCR detection system and CFX Manager Software. Each reaction reagent (20 μL) contained 0.4 μL forward and reverse primers (10 μM), 2 μL cDNA (200±5 ng/μL), 10 μL SYBR Green Supermix (Vazyme, Nanjing, China) and 6.2 μL double-distilled water. Reaction was performed by a two-step method: 95 ℃ for 5 min; 40 cycles of 95 ℃ for 10 s; and 60 ℃ for 30 s. The glyceraldehyde-3-phosphate dehydrogenase (GAPDH) gene was used for internal normalization in each RT-PCR, and the 2−ΔΔCT method (43) was used to calculate the relative gene expression. All samples were performed with three replicates.

Multiple sequence alignment

MUSCLE was used to generate the protein multiple sequence alignment of amino acid decarboxylases with default settings (44). ESPript 3.x was used to display the multiple sequence alignment (45).

Phylogenetic analysis

The CsAlaDC was used as a query for searching its homologs in Embryophyta using BLASTP (46) (version 2.13.0). To fetch homologs of CsAlaDC in distant species, we performed a two-round blast. We first search for the best hit in the given order under Embryophyta. Then we blast each species under a provided order using the order best hit as a query and as many as 10 hits were kept. The sequences with e values <= 0.001 were removed. We further dropped sequences with lengths less than 200 or larger than 800 amino acids. Then multiple sequence alignment was performed via clustalo (version 1.2.4) with default settings on these filtered sequences (47). The NJ algorithm implemented in clustalw (version 2.1) was adopted to build the phylogenetic tree (48). The R package ggtree (version 3.2.1) and universalmotif (version 1.12.4) were used for tree and motif visualization (49).

eLife Assessment

This study reports comparative biochemical and structural analysis of two PLP decarboxylase enzymes from plants. The work is useful because of the potential application of these enzymes in industrial theanine production. The results, particularly the x-ray crystal structures, provide a solid basis for understanding substrate specificity. The paper will be of interest to enzymologists studying PLP enzymes and those working on enzyme engineering in plants.

Reviewer #2 (Public Review):

Summary:

The manuscript focuses on comparison of two PLP-dependent enzyme classes that perform amino acyl decarboxylations. The goal of the work is to understand the substrate specificity and factors that influence catalytic rate in an enzyme linked to theanine production in tea plants.

Strengths:

The work includes x-ray crystal structures of modest resolution of the enzymes of interest. These structures provide the basis for design of mutagenesis experiments to test hypotheses about substrate specificity and the factors that control catalytic rate. These ideas are tested via mutagenesis and activity assays, in some cases both in vitro and in plants.

Weaknesses:

Although improved in a revision, the manuscript could be more clear in explaining the contents of the x-ray structures and how the complexes studied relate to the reactant and product complexes. The manuscript could also be more concise, with a discussion section that is largely redundant with the results and lacking in providing scholarly context from the literature to help the reader understand how the current findings fit in with work to characterize other PLP-dependent enzymes or protein engineering efforts. Some of the figures lack sufficient clarity and description. Some of the claims about the health benefits of tea are not well supported by literature citations.

Reviewer #3 (Public Review):

In the manuscript titled "Structure and Evolution of Alanine/Serine Decarboxylases and the Engineering of Theanine Production," Wang et al. solved and compared the crystal structures of Alanine Decarboxylase (AlaDC) from Camellia sinensis and Serine Decarboxylase (SerDC) from Arabidopsis thaliana. Based on this structural information, the authors conducted both in vitro and in vivo functional studies to compare enzyme activities using site-directed mutagenesis and subsequent evolutionary analyses. This research has the potential to enhance our understanding of amino acid decarboxylase evolution and the biosynthetic pathway of the plant specialized metabolite theanine, as well as to further its potential applications in the tea industry.

Author response:

The following is the authors’ response to the previous reviews.

Response to reviewer’s comments

Reviewer #2 (Public Review):

Summary:

The manuscript focuses on comparison of two PLP-dependent enzyme classes that perform amino acyl decarboxylations. The goal of the work is to understand the substrate specificity and factors that influence catalytic rate in an enzyme linked to theanine production in tea plants.

Strengths:

The work includes x-ray crystal structures of modest resolution of the enzymes of interest. These structures provide the basis for design of mutagenesis experiments to test hypotheses about substrate specificity and the factors that control catalytic rate. These ideas are tested via mutagenesis and activity assays, in some cases both in vitro and in plants.

Weaknesses:

Although improved in a revision, the manuscript could be more clear in explaining the contents of the x-ray structures and how the complexes studied relate to the reactant and product complexes. The manuscript could also be more concise, with a discussion section that is largely redundant with the results and lacking in providing scholarly context from the literature to help the reader understand how the current findings fit in with work to characterize other PLP-dependent enzymes or protein engineering efforts. Some of the figures lack sufficient clarity and description. Some of the claims about the health benefits of tea are not well supported by literature citations.

Thank you for your insightful comments on our manuscript and your recognition of the strengths of our study. We understand your concerns about the weaknesses mentioned, and we have addressed them appropriately in the revised manuscript. We acknowledge that the discussion section needs to be improved for conciseness and context. We have revised this part by removing the redundant content. We also acknowledge your comments concerning the clarity and description of some figures. We have revisited these figures and revised them, ensuring they are clear and adequately described. Lastly, concerning the claims about the health benefits of tea, we understand your concern about the lack of supporting citations. We ensure to back such claims with valid literature or, if necessary, omit these statements.

Reviewer #2 (Recommendations For The Authors):

(1) Line 21: Alanine Decarboxylase should not be capitalized.

Thank you very much for your careful reading of the manuscript. We have corrected it in the revised manuscript.

(2) Line 31: Grammatical error. Also not clear what "evolution analysis" means here. Revise to "Structural comparisons led us to..."

Thank you very much for your careful reading of the manuscript. We have corrected it in the revised manuscript.

(3) Line 34: Revise to "Combining a double mutant of CsAlaDC"

Thank you very much for your careful reading of the manuscript. We have corrected it in the revised manuscript.

(4) Line 35: Change word order to "increased theanine production 672%"

Thank you very much for your careful reading of the manuscript. We have corrected it in the revised manuscript.

(5) Line 37: meaning unclear. Revise to "provides a route to more efficient biosynthesis of theanine."

Thank you very much for your careful reading of the manuscript. We have corrected it in the revised manuscript.

(6) Line 44: I'm not sure that the "health effects" of tea have been proven in placebo controlled studies. And the references provided (2-4 and 5) do not describe original research articles supporting these claims. I would suggest removing these statements from the introduction and at later points in the manuscript.

Thank you for your thoughtful feedback and suggestions. Based on your suggestion, we have removed these statements: "The popularity of tea is determined by its favorable flavor and numerous health benefits (2-4). The flavor and health-beneficial effects of tea are conferred by the abundant secondary metabolites, including catechins, caffeine, theanine, volatiles, etc (5). " As for the subsequent statement: " It has also many health-promoting functions, including neuroprotective effects, enhancement of immune functions, and potential anti-obesity capabilities, among others. " the referenced literature cited can substantiate this conclusion.

(7) Line 58: insert "the" between provided and basis

Thank you very much for your careful reading of the manuscript. We have corrected it in the revised manuscript.

(8) Line 100: Not clear what this phrase means, "As expected, CsSerDC was closer to AtSerDC" Please clarify - closer to what?

We apologize for any confusion caused by the unclear phrasing. When referring to "CsSerDC was closer to AtSerDC," we intended to convey that CsSerDC exhibits a higher degree of sequence homology with AtSerDC than it does with the other enzymes evaluated in our investigation. However, a 1.29% difference between 86.21% and 84.92% in amino acid similarity is not statistically significant (Figure 1B and Supplementary table 1 in the original manuscript), we have deleted the relevant descriptions in the revised manuscript.

(9) Line 112: "were constructed into" makes no sense. It would be better to say the genes for the proteins of interest were inserted into the overexpression plasmid.

Thank you very much for your careful reading of the manuscript. We have corrected it in the revised manuscript.

(10) Line 115: missing the word "the" between generated and recombinant

Thank you very much for your careful reading of the manuscript. We have corrected it in the revised manuscript.

(11) Line 121: catalyze not catalyzed

Thank you very much for your careful reading of the manuscript. We have corrected it in the revised manuscript.

(12) Lines 129 and 130: The reported Km values are really large - in the mM range. Do these values make sense in terms of the available concentrations of the substrates inside the cell?

The content of alanine in tea plant roots ranges from 0.28 to 4.18 mg/g DW (Yu et al., 2021; Cheng et al., 2017). Correspondingly, the physiological concentration of alanine is 3.14 mM to 46.92 mM, in tea plant roots. The content of serine in plants ranges from 0.014 to 17.6 mg/g DW (Kumar et al., 2017). Correspondingly, the physiological concentration of serine is 0.13 mM to 167.48 mM in plants. Therefore, in this study, the Km values are within the range of available substrate concentrations inside the cell.

Yu, Y. et al. (2021) Glutamine synthetases play a vital role in high accumulation of theanine in tender shoots of albino tea germplasm "Huabai 1". J. Agric. Food Chem. 69 (46),13904-13915.

Cheng, S. et al. (2017) Studies on the biochemical formation pathway of the amino acid L-theanine in tea (Camellia sinensis) and other plants. J. Agric. Food Chem. 65 (33), 7210-7216.

Kumar, V. et al. (2017) Differential distribution of amino acids in plants. Amino Acids. 49(5), 821-869.

(13) Line 211: it is unclear what the phrase "as opposed to wild-type" means. Please clarify.

Thank you very much for your careful reading of the manuscript and valuable suggestions. We intend to communicate that the wild-type CsAlaDC and AtSerDC demonstrate decarboxylase activity, while the mutated proteins have experienced a loss of decarboxylation activity. We have already modified this concern in the revised version of the manuscript.

(14) Line 222: residues not residue

Thank you very much for your careful reading of the manuscript. We have corrected it in the revised manuscript.

(15) Line 227 and Figure 4B: It is not clear what the different sequence logos mean in this part of the figure. The caption is too brief and not helpful. And the sentences describing this figure panel are also not sufficiently clear.

Thank you very much for your careful reading of the manuscript and valuable suggestions. We have provided a more detailed explanation of this section in the revised manuscript and added additional annotations in the figure caption to provide further clarity.

(16) Lines 233 and 234: "in the substrate specificity" is awkwardly worded. I would revise to "in selective binding of the appropriate substrate."

Thank you very much for your careful reading of the manuscript and valuable suggestions. We have meticulously revised the description of this section.

(17) Line 243: a word is missing in this sentence - but I can't figure out the intended meaning or what the missing word is. Rephrase to improve clarity.

Thank you very much for your careful reading of the manuscript and valuable suggestions. We have revised this sentence to: " These findings indicate the essential role of Phe106 in the selective binding of alanine for CsAlaDC. "

(18) Line 255: The "expression system...was carried out" is not correct. I would say the expression system was used - but you probably also want to rearrange the sentences to more directly say what it was used for. Later, the word "the" is also missing.

Thank you very much for your careful reading of the manuscript and valuable suggestions. We have revised this sentence to: "To further verify that Phe106 of CsAlaDC and Tyr111 of AtSerDC were key amino acid residues determining its substrate recognition in planta, we employed the Nicotiana benthamiana transient expression system. "

(19) Line 273: use "understand" instead of "elucidate" and instead of "we proposed a prediction test:" say "we designed a test of the prediction that..."

Thank you very much for your careful reading of the manuscript. We have revised this sentence to: “In light of this observation, we postulated a hypothesis:”

(20) Line 301: I don't think "effectuate" is a word. Replace with something else.

Thank you very much for your careful reading of the manuscript. We have revised the sentence as: " The biosynthetic pathway of theanine in tea plants comprises two consecutive enzymatic steps: alanine decarboxylase facilitates the decarboxylation of alanine to generate EA, while theanine synthetase catalyzes the condensation reaction between EA and Glu to synthesize theanine. "

(21) Line 307: replace "activity" with "ability"

Thank you very much for your careful reading of the manuscript. We have corrected it in the revised manuscript.

(22) Line 322: I didn't find the discussion very useful. Much of it is simply a recap of the results - which is not necessary. The structural comparisons are overly descriptive without providing appropriate rationale or topic sentence structure so that the reader understands why certain details are emphasized. I think the manuscript would be much stronger if this section were not included or integreted more concisely into the results section where appropriate.

Thank you for your constructive comments. We understand your concerns about the discussion section of our manuscript. We acknowledge that the discussion section has redundancies with the result. In response to this, we have revised this section to eliminate unnecessary repetition of the results.

(23) Line 369: "an amino acid devoid of the hydroxyl moiety present in Lys" - what does this mean? Lys does not have a hydroxyl functional group. Please correct so that the sentence makes sense.

Thank you very much for your careful reading of the manuscript. This sentence states that the amino acid occupying the corresponding position in CsAlaDC is Phe, which lacks one hydroxyl functional group as compared to Lys. We have made modifications to the sentence as follows: "In contrast, the equivalent position in CsAlaDC is occupied by Phe, an amino acid lacking the hydroxyl group. This substitution enhances the hydrophobic nature of the substrate-binding pocket. "

(24) Line 370: "This structural nuance portends a predisposition for CsAlaDC to select the comparatively hydrophobic amino acid alanine as its suitable substrate." This sentence also makes no sense - please revise to use simpler language so the meaning is more clear.

Thank you very much for your careful reading of the manuscript and valuable suggestions. We have revised the sentence as follows: " Consequently, CsAlaDC demonstrates a unique predilection, selectively binding Ala (an amino acid with comparatively hydrophobic properties) as its preferred substrate."

(25) Lines 376-384: This section makes several references to "catalytic rings." I have no idea what this term means? If the authors mean a loop structure in the enzyme - please use the term "loop"

Thank you very much for your careful reading of the manuscript and valuable suggestions. We have corrected it in the revised manuscript.

(26) Line 396-397: The authors reference data that is not shown in the manuscript. Either show the data in the results section or do not mention.

Thank you for your insightful comment regarding the unshown data referenced in the manuscript. We have included Supplementary figure 9 in the revised manuscript to display this data.

(27) Line 445-446: what is "mutation technology" - if the authors mean site-directed mutagenesis - please use the simpler and more recognizable terminology.

Thank you very much for your careful reading of the manuscript and valuable suggestions. We have revised the sentence as follows: "Based on the findings of this study, site-directed mutagenesis can be employed to modify enzymes involved in theanine synthesis. This modification enhances the capacity of bacteria, yeast, model plants, and other organisms to synthesize theanine, thereby facilitating its application in industrial theanine production."

Reviewer #3 (Public Review):

In the manuscript titled "Structure and Evolution of Alanine/Serine Decarboxylases and the Engineering of Theanine Production," Wang et al. solved and compared the crystal structures of Alanine Decarboxylase (AlaDC) from Camellia sinensis and Serine Decarboxylase (SerDC) from Arabidopsis thaliana. Based on this structural information, the authors conducted both in vitro and in vivo functional studies to compare enzyme activities using site-directed mutagenesis and subsequent evolutionary analyses. This research has the potential to enhance our understanding of amino acid decarboxylase evolution and the biosynthetic pathway of the plant specialized metabolite theanine, as well as to further its potential applications in the tea industry.

Thank you very much for taking the time to review this manuscript. We appreciate all your insightful comments.

Reviewer #3 (Recommendations For The Authors):

The additional material added by the authors addresses some of the previously raised questions and enhances the manuscript's quality. However, certain critical issues we pointed out earlier remain unaddressed. Some of the new data also raises new questions. To provide readers with more comprehensive data, the authors should include additional quantitative data and convert the data presented in the reviewer's comments into supplemental figure format.

Thank you for acknowledging the improvements in the revised manuscript and providing further valuable feedback. We understand your concern about the critical issues that have not been fully addressed and the new questions raised by some of the newly added data. We have strived to address these issues with additional analysis and clarification in our subsequent revision. Regarding your suggestion for more quantitative data and converting the data mentioned in the reviewer's comments into a supplemental figure format, we agree that this would provide a more comprehensive view of the results. We have reformatted the relevant data into supplemental figures to enhance the clarity and accessibility of information. We are grateful for the time and effort you have dedicated to improving our manuscript.

* Page 5 & Figure 1B

"As expected, CsSerDC was most closed to AtSerDC, which implies that they shared similar functions. However, CsAlaDC is relatively distant from CsSerDC."

: In Figure 1B, CsSerDC and AtSerDC are in different clades, and this figure does not show that the two enzymes are closest. To provide another quantitative comparison, please provide a matrix table showing amino acid sequence similarities as a supplemental table.

Comment: I don't believe that a 1.29% difference between 86.21% and 84.92% in amino acid similarity is statistically significant. Although the authors have rephrased the original sentence, it's improbable that this small 1.29% difference can explain the observed distinction.

Many thanks. We have carefully considered your comments. Indeed, the 1.29% difference in amino acid similarity cannot reflect the functional difference between the AlaDC and SerDC proteins. We have deleted the relevant descriptions in the revised manuscript.

* Page 6, Figure 2, Page 23 (Methods)

"The supernatants were purified with a Ni-Agarose resin column followed by size-exclusion chromatography."

: What kind of SEC column did the authors use? Can the authors provide the SEC elution profile comparison results and size standard curve?

Comment: The authors should include the SEC elution profiles as a supplemental figure or incorporate them as a panel in Figure 2. Furthermore, they should provide a description of the oligomeric state of each protein in this experiment. Additionally, there is a significant difference between CsSerDC (65.38 mL) and CsAlaDC (74.37 mL) elution volumes. Can this difference be explained structurally? In comparison to the standard curve of molecular weight provided by the authors, it appears that these proteins are at least homo-tetramers, which contradicts the description in the text. This should be re-evaluated and clarified.

Thank you very much for your careful reading of the manuscript and valuable suggestions. We have included the SEC elution profile in Supplemental figure 1A and added descriptions of the oligomeric states of proteins in the revised manuscript. CsSerDC was eluted at 65.38 mL, corresponding to a molecular weight of 292 kDa, which is five times the monomeric protein (54.7 kDa). However, due to the absence of CsSerDC crystal structure, it remains uncertain whether the protein forms a pentamer. AtSerDC was eluted at 72.25 mL, with a corresponding molecular weight of 155 kDa, which is 3.3 times the monomer (47.3 kDa). CsAlaDC was eluted at 74.37 mL, with a corresponding molecular weight of 127 kDa, which is 2.7 times the monomer (47.3 kDa). The elution profiles suggest that AtSerDC and CsAlaDC potentially exist in homotrimeric form. This observation stands in contradiction to our subsequent findings where the protein manifests in a dimeric structure. A plausible explanation could be the non-ideal spherical shape of the protein. Under such circumstances, the hydrodynamic radius of the protein could supersede its actual size, potentially leading to an overestimation of the molecular weight on the size-exclusion chromatography [ref].

References:

Burgess, R. R. (2018) A brief practical review of size exclusion chromatography: Rules of thumb, limitations, and troubleshooting. Protein Expression and Purification. 150, 81-85.

Erdner J. M., et al. (2006) Size-Exclusion Chromatography Using Deuterated Mobile Phases. Journal of Chromatography A. 1129(1):41–46.

* Page 6 & Page 24 (Methods)

"The 100 μL reaction mixture, containing 20 mM substrate (Ala or Ser), 100 mM potassium phosphate, 0.1 mM PLP, and 0.025 mM purified enzyme, was prepared and incubated at standard conditions (45 {degree sign}C and pH 8.0 for CsAlaDC, 40 {degree sign}C and pH 8.0 for AtSerDC for 30 min)."

(1) The enzymatic activities of CsAldDC and AtSerDC were measured at two different temperatures (45 and 40 {degree sign}C), but their activities were directly compared. Is there a reason for experimenting at different temperatures?

(2) Enzyme activities were measured at temperatures above 40{degree sign}C, which is not a physiologically relevant temperature and may affect the stability or activity of the proteins. At the very least, the authors should provide temperature-dependent protein stability data (e.g., CD spectra analysis) or, if possible, temperature-dependent enzyme activities, to show that their experimental conditions are suitable for studying the activities of these enzymes.

Comment: I appreciate the authors for including temperature-dependent enzyme activity data in their study. However, it remains puzzling that plant enzymes were tested at a physiologically irrelevant temperature of 40 and 45 degrees Celsius. Additionally, it may not be appropriate to directly compare enzyme activity measurements at different temperatures. Furthermore, the data at 45 degrees in panel A appears to be an outlier, which contrasts with the overall trend observed in the graph.

We appreciate your point regarding the testing temperatures for plant enzymes. We fully appreciate the importance of conducting experiments under physiologically relevant conditions. But the intent behind operating at these elevated temperatures was to assess the thermal stability of the enzymes, which can be a valuable characteristic in certain applications, such as industrial production processes, and does not necessarily reflect their physiological conditions. Our findings indicate that CsAlaDC exhibits its peak activity at 45 °C. This result aligns with previously reported data in the literature [Bai, P. et al. (2021) figure 4e], thus bolstering our confidence in the reliability of our experimental outcomes.

Relative activity of CsAlaDC at different temperatures.

* Pages 6-7 & Table 1

(1) Use the correct notation for Km and Vmax. Also, the authors show kinetic parameters and use multiple units (e.g., mmol/L or mM for Km).

(2) When comparing the catalytic efficiency of enzymes, kcat/Km (or Vmax/Km) is generally used. The authors present a comparison of catalytic activity from results to conclusion. A clarification of what results are being compared is needed.

Comment: The authors are still comparing catalytic efficiency solely based on the Vmax values. As previously suggested, it would be advisable to calculate kcat/Km and employ it for comparing catalytic efficiencies. Furthermore, based on the data provided by the authors, I conducted a rough calculation of these catalytic efficiencies and did not observe a significant difference, which contrasts with the authors' statement, "These findings indicated that the catalytic efficiency of CsAlaDC is considerably lower than that of both CsSerDC and AtSerDC." This discrepancy requires clarification.

We want to express our sincere appreciation for your meticulous review and constructive suggestions. We understand the importance of accurately comparing catalytic efficiencies using Kcat/Km values, rather than solely relying on Vmax values. Following your suggestion, we recalculated Kcat/Km to reanalyze our results. The computed Kcat/Km for CsSerDC and AtSerDC are 152.7 s-1 M-1 and 184.6 s-1 M-1, respectively. For CsAlaDC, the calculated Kcat/Km is 55.7 s-1 M-1. Therefore, the catalytic efficiency of CsSerDC and AtSerDC is approximately three times that of CsAlaDC. What we intended to convey was that the Vmax of CsAlaDC is lower than that of CsSerDC and AtSerDC. Our description in the manuscript was not accurate, and we have addressed this in the revised version.

* Pages 9 & 10

"This result suggested this Tyr is required for the catalytic activity of CsAlaDC and AtSerDC."

: The author's results are interesting, but it is recommended to perform the experiments in a specific order. First, experiments should determine whether mutagenesis affects the protein's stability (e.g., CD, as discussed earlier), and second, whether mutagenesis affects ligand binding (e.g., ITC, SPR, etc.), before describing how site-directed mutagenesis alters enzyme activity. In particular, the authors' hypothesis would be much more convincing if they could show that the ligand binding affinity is similar between WT and mutants.

Comments: While it is appreciated that you have included CD and UV-vis absorption spectra data, it would be more beneficial to provide quantitative data to address the previously proposed binding affinity. I also recommend presenting the data mentioned in the reviewer's comments as a supplementary figure for better clarity and reference.

Thank you for your valuable feedback and suggestions. I agree that providing quantitative data would lend more support to our findings and better address the proposed binding affinity.

It is generally acknowledged that proteins complexed with PLP exhibit a yellow hue, and the ligand PLP forms a Schiff base structure with the ε-amino group of a lysine residue in the protein, with maximum absorbance around 420 nm. However, during our protein purification process, we observed that the purified protein retained its yellow coloration, even when PLP wasn't introduced into the purification buffer. Subsequent absorbance measurements revealed that the protein exhibited absorbance within the aforementioned wavelength (420 nm) (the experimental results are shown in the following figures), implying an inherent presence of the PLP ligand within the protein. This could have resulted from binding with PLP during the protein's expression in E. coli. Consequently, due to this inseparability between the protein and the ligand, obtaining quantitative data through experimental means becomes unfeasible.

(A) Absorption Spectra of CsAlaDC (WT) and CsAlaDC (Y336F).

(B) Absorption Spectra of AtSerDC (WT) and AtSerDC (Y341F).

Regarding your suggestion about presenting the data mentioned in the reviewer's comments as a supplementary figure, we agree that it is an excellent idea. We have prepared supplementary figure 7 and supplementary figure 8 accordingly, ensuring that they present the required data.

* Page 10

"The results showed that 5 mM L-DTT reduced the relative activity of CsAlaDC and AtSerDC to 22.0% and 35.2%, respectively"

: The authors primarily use relative activity to compare WT and mutants. Can the authors specify the exact experiments, units, and experimental conditions? Is it Vmax or catalytic efficiency? If so, under what specific experimental conditions?

Response: "However, due to the unknown mechanism of DTT inhibition on protein activity, we have removed this part of the content in the revised manuscript."

Comment: I believe this requires a more comprehensive explanation rather than simply removing it from the text.

Although we have observed that DTT is capable of inhibiting enzyme activity, at present, we are unable to offer a comprehensive explanation for the inhibitory effect of DTT on enzyme activity in terms of its structural and catalytic mechanisms. Further research is required to elucidate the mechanism of action of DTT. It is worth noting, however, that our study does not emphasize investigating the specific inhibitory mechanisms of DTT on enzyme activity. Furthermore, the existing findings do not provide an adequate explanation for the observed phenomenon, leading us to exclude this particular aspect from the content.

* Pages 10-12

: The identification of 'Phe106 in CsAlaDC' and 'Tyr111 in AtSerDC,' along with the subsequent mutagenesis and enzymatic activity assays, is intriguing. However, the current manuscript lacks an explanation and discussion of the underlying reasons for these results. As previously mentioned, it would be helpful to gain insights and analysis from WT-ligand and mutant-ligand binding studies (e.g., ITC, SPR, etc.). Furthermore, the authors' analysis would be more convincing with accompanying structural analysis, such as steric hindrance analysis.

Comment: While it is appreciated that you have included UV-vis absorption spectra data, it would be more beneficial to provide quantitative data to address the previously proposed binding affinity. I also recommend presenting the data mentioned in the reviewer's comments as a supplementary figure for better clarity and reference.

Response: Thank you for your valuable feedback and suggestions. Given that the protein forms a complex with PLP during its expression in E. coli and cannot be dissociated from it, obtaining quantitative data via experimental protocols is rendered impracticable.

(A) Absorption Spectra of CsAlaDC (WT) and CsAlaDC (F106Y).

(B) Absorption Spectra of AtSerDC (WT) and AtSerDC (Y111F).

Mutant proteins and wild-type proteins exhibited absorption bands at 420 nm, suggesting the formation of a Schiff base between PLP and the active-site lysine residue.

Regarding your suggestion about presenting the data mentioned in the reviewer's comments as a supplementary figure, we have prepared supplementary figure 7 and supplementary figure 8 accordingly, ensuring that they present the required data.

Version 3

Abstract

Abstract

Ethylamine (EA), the precursor of theanine biosynthesis, is synthesized from alanine decarboxylation by alanine decarboxylase (AlaDC) in tea plants. AlaDC evolves from serine decarboxylase (SerDC) through neofunctionalization and has lower catalytic activity. However, lacking structure information hinders the understanding of the evolution of substrate specificity and catalytic activity. In this study, we solved the X-ray crystal structures of AlaDC from Camellia sinensis (CsAlaDC) and SerDC from Arabidopsis thaliana (AtSerDC). Tyr341 of AtSerDC or the corresponding Tyr336 of CsAlaDC is essential for their enzymatic activity. Tyr111 of AtSerDC and the corresponding Phe106 of CsAlaDC determine their substrate specificity. Both CsAlaDC and AtSerDC have a distinctive zinc finger and have not been identified in any other Group II PLP-dependent amino acid decarboxylases. Based on the structural comparisons, we conducted mutation screen of CsAlaDC. The results indicated that the mutation of L110F or P114A in the CsAlaDC dimerization interface significantly improved the catalytic activity by 110% and 59%, respectively. Combining a double mutant of CsAlaDCL110F/P114A with theanine synthetase increased theanine production 672% in an in vitro system. This study provides the structural basis for the substrate selectivity and catalytic activity of CsAlaDC and AtSerDC and provides a route to more efficient biosynthesis of theanine.

Introduction

Tea is one of the most popular natural non-alcoholic beverages consumed worldwide. About 2 billion cups of tea are consumed worldwide daily (1). Theanine (γ-glutamylethylamide) is a unique non-protein amino acid and the most abundant free amino acid in tea plants (Camellia sinensis). It accounts for more than 50% of the total free amino acids and approximately 1-2% of the dry weight of the new shoots of tea plants (2). Theanine is the secondary metabolite conferring the umami taste of tea infusion and also balances the astringency and bitterness of tea infusion caused by catechins and caffeine (3). It has also many health-promoting functions, including neuroprotective effects, enhancement of immune functions, and potential anti-obesity capabilities, among others (2, 4–9). Therefore, theanine content is highly correlated with green tea quality (10).

Theanine is synthesized from ethylamine (EA) and glutamate (Glu) by theanine synthetase (11, 12). Importantly, the large amount of theanine biosynthesis is determined by the high availability of EA in tea plants (13). Therefore, the evolution of EA biosynthesis in tea plants provided the basis for theanine biosynthesis and the formation of tea quality. EA is synthesized from alanine decarboxylation by alanine decarboxylase (CsAlaDC) (14) (Figure 1A). Indeed, CsAlaDC expression level and catalytic activity largely determine the theanine accumulation level in tea plants (15). As a novel gene, CsAlaDC had not been reported in other plant species before it was identified in tea plants. Previous studies indicated that AlaDC originated from the serine decarboxylase gene (SerDC) by gene duplication in tea plants (16). However, despite sharing highly conserved amino acid sequences, CsAlaDC and CsSerDC specifically catalyze the decarboxylation of alanine and serine, respectively (16) (Figure 1A and B). In addition, CsAlaDC exhibited a much lower enzymatic activity compared with CsSerDC (16). However, the structural basis and key amino acids underlying the evolution of the substrate specificity and enzymatic activity of CsAlaDC are unknown.

Metabolic pathways and sequence analysis of CsAlaDC and SerDCs in plants.

(A) The decarboxylation of serine and alanine in plants. SerDC, serine decarboxylase; AlaDC, alanine decarboxylase. (B) Multiple alignment of the amino acid sequences of AlaDC and SerDCs. The amino acid sequences of the 6 SerDCs from kiwifruit (Actinidia chinensis), grape (Vitis vinifera), coffee (Coffea eugenioides), cocoa (Theobroma cacao) and Arabidopsis. Primary (100%), secondary (80%), and tertiary (60%) conserved percent of similar amino acid residues were shaded in deep blue, light blue and cheer red, respectively. “*” indicated amino acid residues mutated only in CsAlaDC.

SerDC belongs to the Group II pyridoxal-5-phosphate (PLP)-dependent decarboxylase superfamily (17). These Group II amino acid decarboxylases produce many bioactive secondary metabolites and signaling molecules (18–20). In plants, SerDC catalyzes the biosynthesis of ethanolamine from serine, which was first functionally characterized in Arabidopsis thaliana (21) (Figure 1A). Ethanolamine is an important metabolite in the synthesis of phosphatidylethanolamine and phosphatidylcholine, two main phospholipids involved in maintaining eukaryotic membrane structures in eukaryotic cell membranes (22–24). Furthermore, the study has shown that SerDC is essential in the embryogenesis of Arabidopsis (25). SerDC is of vital importance for plant growth and development, but the mechanism of substrate recognition and catalytic activity of SerDC also remains unclear.

In this work, we attempt to understand the mechanism of functional divergence of plant AlaDC and SerDC by analyzing their structural characteristics. Here, we obtained the X-ray crystal structure of CsAlaDC and AtSerDC. According to the crystal structure, we found a distinctive zinc finger structure located at both CsAlaDC and AtSerDC, which has not been identified in any other Group II PLP-dependent amino acid decarboxylases that have been previously characterized. By comparing the substrate binding pockets, we identified Phe106 of CsAlaDC and Tyr111 of AtSerDC as the crucial sites for substrate specificity. By conducting mutation screening based on the protein structures, we identified the amino acids repressing the catalytic activity and discovered that CsAlaDCL110F/P114A exhibited a 2.3-fold increase in catalytic activity compared to that of CsAlaDC and enhanced the engineering of theanine production in vitro.

Results

Enzymatic properties of CsAlaDC, AtSerDC, and CsSerDC

CsAlaDC originates from CsSerDC by gene duplication and neofunctionalization in tea plants, but they catalyze different metabolic processes (Figure 1A). We performed a multiple sequence alignment of the amino acid sequences of the 6 SerDCs from kiwifruit (Actinidia chinensis), grape (Vitis vinifera), coffee (Coffea eugenioides), cocoa (Theobroma cacao), Arabidopsis and tea plants and CsAlaDC (Figure 1B). The results showed that the amino acid sequences of CsAlaDC and SerDCs were highly conserved, but CsAlaDC has amino acid mutations in highly conserved regions compared with SerDCs.

Next, to verify the substrate specificity and enzyme activity, we conducted enzyme activity detection and enzyme kinetics analysis for CsAlaDC, CsSerDC and AtSerDC. The 5′ truncated CsAlaDC and AtSerDC were inserted into the pET22b and pET28a expression vectors to generate recombinant plasmids pET22b-CsAlaDC and pET28a-AtSerDC, correspondingly. Additionally, the full-length protein of CsSerDC was inserted into the pET28a expression vector to generate the recombinant plasmid pET28a-CsSerDC. Subsequently, the recombinant proteins were expressed in Escherichia coli and purified via nickel affinity chromatography. The purified proteins were examined through SDS-PAGE analysis (Figure 2A), supplemented by size-exclusion chromatography assessments (Supplementary figure 1A), it seems that the protein manifests in an oligomeric configuration within the solution.

Purification and characterization of CsAlaDC, AtSerDC, and CsSerDC.

(A) Identification of CsAlaDC, AtSerDC, and CsSerDC using SDS-PAGE. (B) Detection of enzyme activities of CsAlaDC, AtSerDC, and CsSerDC by UPLC. (C) Reaction rates of substrates with different concentrations catalyzed by CsAlaDC, AtSerDC, and CsSerDC.

To verify the substrate specificity of CsAlaDC, AtSerDC, and CsSerDC, enzyme activity assays were conducted using Ala and Ser as substrates, followed by product identification via UPLC analysis. The results showed that CsAlaDC can effectively catalyze the decarboxylation of alanine to generate EA, whereas its catalytic efficacy towards serine was relatively inferior. Conversely, AtSerDC and CsSerDC selectively catalyzed serine decarboxylation to yield ethanolamine but did not exhibit activity towards alanine (Figure 2B). These findings confirmed that CsAlaDC and SerDCs do not have promiscuous decarboxylase activity on Ala and Ser.

The kinetic properties of CsAlaDC, AtSerDC, and CsSerDC were determined through the use of corresponding substrates (Figure 2C), and the results are presented in Table 1. The kinetic parameter Km of CsAlaDC was determined to be 1.215 mM, which is similar to that of AtSerDC and CsSerDC at 1.522 mM and 2.364 mM, respectively. However, AtSerDC has a Vmax of 7.053 μmo L-1 s-1, which is 4-fold that of CsAlaDC’s Vmax of 1.709 μmo L-1 s-1; while CsSerDC has a Vmax of 9.031 μmo L-1 s-1, which is 5.3-fold that of CsAlaDC. The findings suggest that the enzymatic catalytic efficiency of CsSerDC and AtSerDC is approximately triple compared to that of CsAlaDC, and the Vmax of CsAlaDC is considerably lower than that of both CsSerDC and AtSerDC. The observed disparity in Vmax suggests that CsAlaDC and SerDCs may exhibit discrepancies in substrate-to-product conversion rates under saturated substrate conditions. It is plausible that SerDCs possess a higher turnover rate, facilitating the rapid conversion of substrate to product and subsequent release of the enzyme. Another possibility is that SerDCs demonstrate enhanced stability, thereby maintaining superior catalytic activity even at higher substrate concentrations, consequently leading to a higher Vmax value. Due to the similarity in catalytic activity between CsSerDC and AtSerDC, we chose the more representative AtSerDC for further analysis.

The overall structures of CsAlaDC and AtSerDC

To enhance our comprehension of CsAlaDC and AtSerDC, we conducted structural analyses of these two proteins. Through optimization of crystallization conditions, we successfully determined the crystal structures of CsAlaDC, CsAlaDC-EA, and AtSerDC at resolutions of 2.50 Å, 2.60 Å and 2.85 Å, respectively (Supplementary table 1).

The overall structure of CsAlaDC or AtSerDC is homodimeric with two subunits exhibiting an asymmetrical arrangement (Figure 3A and 3D). The monomer of CsAlaDC or AtSerDC is divided into three distinct structural domains: an N-terminal domain (N-terminal –104 aa in CsAlaDC, N-terminal –109 aa in AtSerDC), a large domain (105-364 aa in CsAlaDC, 110-369 aa in AtSerDC), and a small C-terminal domain (365 aa-C-terminus in CsAlaDC, 370-C-terminus as in AtSerDC), colored light pink, khaki, and sky blue, respectively. Compared to the full-length protein, the N-terminal structures of CsAlaDC and AtSerDC were truncated by 60 and 65 amino acid residues, respectively. The N-terminal domain of the truncated protein contains a long α-helix, which is connected to the large domain through a long loop. The helix of one subunit is antiparallel to the corresponding helix in the neighboring subunit, forming a clamp. Therefore, the N-terminal domain may not be an independent folding unit and is likely only stable in dimers. The large domain contains a seven-stranded mixed β-sheet surrounded by eight α-helices, while the C-terminal domain is composed of a three-stranded antiparallel β-sheet and three α-helices (Supplementary figure 1). The catalytic site of the enzyme is positioned within a superficial crevice at the junction of two subunits forming a dimeric arrangement. The amino acid residues derived from both subunits participate in the binding of the PLP, effectively anchoring it to the active site. In the catalytic site, one monomer accommodates PLP, while the other monomer also contributes to PLP binding and enzymatic function (Figure 3A and 3D).

Crystal structures of CsAlaDC and AtSerDC.

(A) Dimer structure of CsAlaDC. The color display of the N-terminal domain, large domain, and C-terminal domains of chain A is shown in light pink, khaki and sky blue, respectively. Chain B is shown in spring green. The PLP molecule is shown as a sphere model. The zinc finger structure at the C-terminus of CsAlaDC is indicated by the red box. The gray spheres represent zinc ions, while the red dotted line depicts the coordination bonds formed by zinc ions with cysteine and histidine. (B) The 2Fo-Fc electron density maps of K309-PLP-EA (contoured at 1σ level). The PLP is shown in violet, the K309 is shown in spring green, and the EA is shown in lightblue. (C) Active center of the CsAlaDC-EA complex, with hydrogen bonds denoted by black dotted lines. “*” denotes the amino acids on adjacent subunits. (D) Dimer structure of AtSerDC. The color display of the N-terminal domain, large domain, and C-terminal domains of chain A is shown in light pink, khaki and sky blue, respectively. Chain B is shown in cyan. The PLP molecule is shown as a sphere model. The zinc finger structure at the C-terminus of AtSerDC is indicated by the red box. The gray spheres represent zinc ions, while the red dotted line depicts the coordination bonds formed by zinc ions with cysteine and histidine. (E) Active center of the AtSerDC, with hydrogen bonds denoted by black dotted lines. “*” denotes the amino acids on adjacent subunits. (F) The monomers of CsAlaDC and AtSerDC are superimposed. CsAlaDC is depicted in spring green, while AtSerDC is shown in plum. The conserved amino acid catalytic loop is indicated by the red box. (G) Amino acid residues of the active center in CsAlaDC apo and CsAlaDC-EA complex are superimposed. CsAlaDC apo is shown in floral white, while CsAlaDC-EA complex is shown in spring green. (H) The relative activity of wild-type CsAlaDC and its Y336F mutant (left), as well as wild-type AtSerDC and its Y341F mutant (right) is shown.

Notably, our investigation has revealed the presence of a distinctive zinc finger structure (as depicted in Figure 3A and 3D) located at the C-terminus of CsAlaDC and AtSerDC. This structure is composed of a loop structure spanning 17 amino acid residues, wherein coordination of Zn2+ is facilitated by three Cys residues and one His residue. Importantly, this particular configuration is exclusive to the two proteins under examination and has not been identified in any other Group II PLP-dependent amino acid decarboxylases that have been previously characterized (26, 27).

The crystal structures of CsAlaDC-EA complex and AtSerDC were obtained and further analyzed. In the former complex, PLP bound to Lys309 via a Schiff base linkage (internal aldimine form), and the pyridine moiety of PLP is positioned between the imidazole ring of His196 and the methyl group of Ala279 in a parallel orientation to the imidazole ring. The carboxylic group of Asp277 stabilizes the N1 of PLP by a salt bridge interaction, providing the latter with the strong electrophilicity necessary to stabilize carbon ion intermediates during enzymatic catalysis. The O3 forms hydrogen bonds with Thr247 and Lys309. Additionally, His308, Gly169, Thr170, Lys309, and Ser347*(“*” denotes the amino acids on adjacent subunits) establish a hydrogen bonding network with the phosphate group of PLP, consequently fortifying its attachment to the protein (Figure 3C). Similarly, in the latter structure, the pyridine ring of PLP was sandwiched between the imidazole ring of His201 and the methyl group of Ala284. The N1 of PLP formed a salt bridge with Asp282, while the O3 formed hydrogen bonds with Thr252. The phosphate group of PLP established hydrogen bonding interactions with Lys314, His313, Gly174, Thr175, and Ser352* (Figure 3E). The aforementioned amino acid residues are observed across various Group II PLP-dependent amino acid decarboxylases and exhibit considerable stability (see further discussion below), suggesting that they could share common features in terms of catalytic mechanisms.

Factors affecting CsAlaDC and AtSerDC activity

The Group II PLP-dependent amino acid decarboxylases are dimeric enzymes with each monomer containing an active site located within a shallow cavity (28). A flexible loop of amino acids originating from one monomer extends into the active site of the other monomer and plays a crucial role in catalysis. The catalytic loop in AtSerDC, consisting of amino acid residues 328-341, is well-structured and exhibits clear electron density. However, the corresponding loop in CsAlaDC is disordered (Figure 3F). The aforementioned loop harbors a conserved Tyr residue that is believed to function as a proton donor of the carbanion in the catalytic process (29). In particular, CsAlaDC Tyr336 and AtSerDC Tyr341 correspond to that Tyr residue (Supplementary figure 2). Upon superimposing the active site residues of CsAlaDC apo and CsAlaDC-EA complex, it was observed that the hydroxyl group of Tyr336* in the CsAlaDC-EA complex exhibits a substantial 60-degree deviation along the Cα-Cβ bond (Figure 3G). To assess the function of the Tyr, we substituted the corresponding Tyr residues in CsAlaDC and AtSerDC with Phe. The resulting CsAlaDCY336F and AtSerDCY341F mutants were then exposed to Ala and Ser, respectively. We measured the production of EA or ethanolamine in the reaction mixtures. Our findings indicate that these mutants catalyze abortive decarboxylation, as evidenced by the absence of detectable EA and only a small amount of ethanolamine observed in the reaction mixture (Figure 3H). This result suggested this Tyr is required for the catalytic activity of CsAlaDC and AtSerDC.

Identification of key amino acids for the substrate specificity

The superposition of amino acid residues in the substrate binding pocket of CsAlaDC and AtSerDC revealed that all residues are identical, except for the substitution of Tyr at position 111 in AtSerDC with Phe at position 106 in CsAlaDC (Figure 4A). This observation suggests a potential role for this specific residue in determining the selective binding of the appropriate substrate.

Key amino acid residues for substrate recognition.

(A) Superposition of substrate binding pocket amino acid residues in CsAlaDC and AtSerDC. The amino acid residues of CsAlaDC are shown in spring green, the amino acid residues of AtSerDC are shown in plum, with the substrate specificity-related amino acid residue highlighted in a red ellipse. (B) Active-site-lining amino acid residues of SDC homologs from Embryophyta were identified. The height of each amino acid is scaled proportionally to the amount of information content (measured in bits). The first line depicts the conserved motif in all SerDC homologs from Embryophyta, whereas lines 2-5 represent the conserved motifs based on the variable third amino acid residue. (C) Histogram showing the distribution of the number of key motifs. (D) Histogram showing the number of key motifs in different plant orders. (E) Relative enzyme activities of wild-type CsAlaDC and mutant protein CsAlaDCF106Y against Ala substrate (columns 1 and 2), and enzyme activities of wild-type AtSerDC and various AtSerDC mutant proteins against Ala substrate (columns 3-9) are presented. The percentage graph shows the relative activity of each protein compared to wild-type CsAlaDC activity (taken as a 100% benchmark). (F) Relative enzyme activities of wild-type AtSerDC and AtSerDC mutant proteins (columns 1-7) against Ser substrates, and enzyme activities of wild-type CsAlaDC and mutant protein CsAlaDCF106Y against Ser substrates(columns 8, 9) were measured. The percentage graph shows the relative activity of each protein compared to the wild-type AtSerDC (taken as a 100% benchmark). Three independent experiments were conducted. (G) The EA contents of AtSerDC and its mutant AtSerDCY111F in N. benthamiana. (H) The EA contents of CsAlaDC and its mutant CsAlaDCF106Y in N. benthamiana. The significance of the difference (P<0.05) was labeled with different letters according to Duncan’s multiple range test.

To gain further insights into the functional role of this residue (Tyr111 in AtSerDC or Phe106 in CsAlaDC) in selective binding of the appropriate substrate, we identified 563 potential serine decarboxylases in Embryophyta using the amino acid sequences of AtSerDC (Supplementary figure 3A). By comparing the amino acid sequences of these serine decarboxylases, we observed that the corresponding residues to Tyr111 or Phe106 are located within a conserved motif comprising 9 amino acids (Figure 4B). Within this motif, the first two residues, Y (Tyrosine) and P (Proline) are completely conserved across all 563 proteins. However, the third residue where Tyr111 in AtSerDC or Phe106 in CsAlaDC is positioned exhibits variability among Y, T (Threonine), F (Phenylalanine), V (Valine), A (Alanine), I (Isoleucine), L (Leucine), or other amino acids (Figure 4B). Remarkably similar to Tyr111, this residue is predominantly Y in 83.7% of these homologs; conversely, F was found to be present in only 2.3% of these homologs within plant species belonging to Poales, Asterales, Ranunculales, Ericales, Caryophyllales and Nymphaeales orders (Figure 4C-4D).

To verify the role of Tyr111 in AtSerDC or Phe106 in CsAlaDC in the selective binding of the appropriate substrate, we introduced mutations in Tyr111 to Phe (AtSerDCY111F) and Phe106 to Tyr (CsAlaDCF106Y), followed by in vitro enzymatic activity assays. The results revealed that AtSerDCY111F acquired alanine decarboxylase activity, whereas CsAlaDCF106Y completely lost its alanine decarboxylase activity (Figure 4E). Additionally, we generated other mutations of Tyr111 in AtSerDC, including AtSerDCY111A, AtSerDCY111I, AtSerDCY111L, AtSerDCY111V and AtSerDCY111W. Among these mutants, only AtSerDCY111I exhibited a marginal level of alanine decarboxylase activity (Figure 4E). These findings indicate the essential role of Phe106 in the selective binding of alanine for CsAlaDC.

On the other side, both CsAlaDC and CsAlaDCF106Y exhibit a very low level of serine decarboxylase activity (Figure 4F). Moreover, compared with AtSerDC, the serine decarboxylase activity of AtSerDCY111F was about 30% of AtSerDC; AtSerDCY111I retained lower than 5% of serine decarboxylase activity of AtSerDC; while other mutations, including AtSerDCY111A, AtSerDCY111L, AtSerDCY111V and AtSerDCY111W abolished the serine decarboxylase activity (Figure 4F). These results suggested the Tyr111 of AtSerDC is also important for the selective binding of serine for AtSerDC.

To further verify that Phe106 of CsAlaDC and Tyr111 of AtSerDC were key amino acid residues determining the selective binding of substrates in planta, we employed the Nicotiana benthamiana transient expression system. To this end, A. tumefaciens strain GV3101 (pSoup-p19), carrying recombinant plasmid, was infiltrated into leaves of 5-week-old N. benthamiana plants, and the pCAMBIA1305 empty vector was used as the control (EV). Relative mRNA levels were detected (Supplementary figure 4), indicating that they had been successfully overexpressed in tobacco. Here, we found that a high level of EA was produced in the CsAlaDC-expressing tobacco leaves; while no EA product was detected in tobacco leaves infiltrated with the mutant CsAlaDCF106Y. In addition, we did not detect EA products in the AtSerDC-expressing tobacco leaves, while a high level of EA was detected in tobacco leaves infiltrated with mutant AtSerDCY111F. As anticipated, EA was not detected in tobacco leaves infiltrated with EV (Figure 4G and H). These results further verified the critical role of Phe106 in the selective binding of alanine for CsAlaDC.

Key amino acids for the evolution of CsAlaDC enzymatic activity

CsAlaDC and AtSerDC have a high sequence similarity of 74.5% and a nearly identical structure with an RMSD (root mean square deviation) of only 0.77Å for monomer structures. However, the two enzymes catalyze different amino acid decarboxylation reactions, and AtSerDC exhibits significantly higher Vmax than CsAlaDC (Figure 2B and C). In light of this observation, we postulated a hypothesis: EA, generated via Ala decarboxylation by CsAlaDC, can be toxic and harmful to plants if accumulated excessively. Thus, during the evolution of plant serine decarboxylase into alanine decarboxylase, the enzyme has evolved not just to alter the substrate preference but also to reduce catalytic activity to control EA production within a suitable range. Based on this hypothesis, we suggest that mutating specific amino acids in CsAlaDC to those corresponding amino acids in SerDCs could enhance its activity, and the results could provide insights into the evolution of CsAlaDC enzymatic activity.

Dimerization is essential to AlaDC/SerDC activities because the active site is composed of residues from two monomers. To identify the amino acids repressing the enzymatic activity of CsAlaDC during evolution from SerDC, we analyzed the crystal structures and the amino acids at the dimer interface between CsAlaDC and AtSerDC. This analysis revealed that the amino acids at the dimer interface at positions 66, 97, 110, 114, 116, 117, 122, 315 and 345 are different in CsAlaDC and SerDCs (Figure 1B). Therefore, we mutated these amino acids of CsAlaDC into those in the corresponding positions of AtSerDC and CsSerDC, and performed enzyme activity assays (Figure 5A).

Mutations enhance CsAlaDC enzyme activity and theanine synthesis in vitro.

(A) Relative enzyme activities of CsAlaDC mutant proteins against Ala substrate. (B) Relative enzyme activities of CsAlaDCL110F, CsAlaDCP114A, and CsAlaDCL110F/P114A against Ala substrate. (C) Histogram showing the relative content of theanine resulting from different combinations of alanine decarboxylase and theanine synthetase. Three independent experiments were conducted.

The results demonstrated that CsAlaDCL110F and CsAlaDCP114A exhibited significantly enhanced enzyme activity compared to the wild-type CsAlaDC, with a 2.1-fold and 1.59-fold increase, respectively. Furthermore, the catalytic activity of the CsAlaDCL110F/P114A double mutant exhibits a remarkable 2.3-fold increase compared to that of the wild-type protein (Figure 5B). These findings suggested a critical role of Leu110 and Pro114 of CsAlaDC in the evolution of enzymatic activity and provided a basis to improve CsAlaDC activity. It is possible that these amino acid residues could potentially augment the hydrophobic nature of the protein dimer interface.

In Vitro synthesis of theanine

The biosynthetic pathway of theanine in tea plants comprises two consecutive enzymatic steps: alanine decarboxylase facilitates the decarboxylation of alanine to generate EA, while theanine synthetase catalyzes the condensation reaction between EA and Glu to synthesize theanine. By conducting mutation screens, we discovered a CsAlaDC mutant protein (L110F/P114A) that exhibited a 2.3-fold higher catalytic activity compared to the wild-type protein. Subsequently, we employed an in vitro theanine synthesis system utilizing CsAlaDC and either glutamine synthetase (PsGS [Pseudomonas syringae pv. syringae]) or gamma-glutamate methylamine ligase (MmGMAS [Methylovorus mays]) which have the ability to synthesize theanine from EA and Glu (30, 31).

The results illustrated the successful synthesis of theanine using CsAlaDC in conjunction with two theanine synthetases, employing Ala and Glu as substrates (Figure 5C). The theanine content generated by the combination of CsAlaDCL110F and MmGMAS, as well as CsAlaDCL110F/P114A and MmGMAS, was 4.57-fold and 6.72-fold higher than the content produced in the combination of wild-type CsAlaDC and MmGMAS, respectively. Similarly, when combined with the PsGS, comparable outcomes are observed as well. The theanine content resulting from the combination of CsAlaDCL110F and PsGS, as well as CsAlaDCL110F/P114A and PsGS, exhibit enhancements of 1.62-fold and 4.33-fold compared to the wild-type protein combination, respectively (Figure 5C). Hence, the utilization of CsAlaDCL110F/P114A could effectively enhance the theanine production yield and thus, holds potential for large-scale engineering production of theanine.

Discussion

The common and distinctive structural characteristics of CsAlaDC and AtSerDC compared with other amino acid decarboxylases

CsAlaDC and AtSerDC are Group II PLP-dependent amino acid decarboxylases, which exhibit numerous structural features common to other amino acid decarboxylases. The Dali search (32) revealed that AtSerDC and CsAlaDC share structural similarities with 7CIG (MetDC [Streptomyces sp. 590]), 7ERV (HisDC1 [Photobacterium phosphoreum]), 6KHO (TrpDC [Oryza sativa Japonica Group]), 4E1O (HisDC2 [Homo sapiens]), 6JY1 (AspDC [Methanocaldococcus jannaschii]), and 5GP4 (GluDC [Levilactobacillus brevis]). The monomers of these amino acid decarboxylases all consist of three characteristic domains: the C-terminal domain, the large domain and the N-terminal domain (Supplementary figure 5). The amino acid residues that bind to PLP cofactors are conserved across multiple enzymes (Supplementary figure 5 and 6). Out of the eight proteins analyzed, a total of thirteen amino acid residues were found to be conserved across all sequences (Glu143, Glu171, Lys201, Gly245, Thr247, Asp253, His275, Asp277, Ala279, Pro286, Ser302, Lys309, and Tyr336 in CsAlaDC). Some of them are situated within the active center and play a role in stabilizing PLP or facilitating catalytic reactions. Conversely, other residues reside outside the active center and their function remains unclear (Supplementary figure 6).

Group II PLP-dependent amino acid decarboxylases are characterized by the existence of a highly flexible loop, which is of great significance for the catalytic mechanism of decarboxylase (33). The loops are located at the dimer interface and extend to the active sites of other monomers in a closed conformation. Prior research has established that the conserved amino acid residue Tyr within the loop plays a crucial role in catalysis by donating protons to carbanions of quinone intermediates that arise following decarboxylation (34, 35). Our experiments demonstrate that the substitution of the corresponding Tyr with Phe in the loop renders the protein inactive. Through the application of circular dichroism, we have corroborated that the stability of the protein is not compromised by mutations (Supplementary figure 7A, B, D, E). Both the mutant and wild-type proteins manifest absorption bands at the 420 nm wavelength, signifying the formation of a Schiff base between PLP and the lysine residues found at the active site. Given that PLP has been incorporated during the protein purification stage, the absorbance observed for both the mutant and wild-type proteins at 420 nm is consistent when compared at equivalent concentrations (Supplementary figure 8). This corroboration implies that the protein mutation does not interfere with the binding to PLP. Consequently, the inactivation observed in the mutant proteins is not a consequence of instability incited by corresponding mutations, nor changes in the binding affinity between the protein and PLP induced by these mutations.

We observed that the substitution of Phe106 for Tyr in CsAlaDC rendered CsAlaDC inactive, while the substitution of Tyr111 for Phe in AtSerDC enabled alanine decarboxylase activity, and the alteration in the activity of the mutant protein bears no correlation to protein stability and affinity for PLP (Supplementary figure 7A, C, D, F and Supplementary figure 8), suggesting that the amino acid in the position is critical for determining substrate specificity. Integrative crystallographic structure of the two enzymes, we speculate the hydrophilic nature of the Lys residue in AtSerDC predisposes the active site to preferentially accommodate the hydrophilic amino acid Ser as its substrate. In contrast, the equivalent position in CsAlaDC is occupied by Phe, an amino acid lacking the hydroxyl group. This substitution enhances the hydrophobic nature of the substrate-binding pocket. Consequently, CsAlaDC demonstrates a unique predilection, selectively binding Ala (an amino acid with comparatively hydrophobic properties) as its preferred substrate.

The monomeric configuration of CsAlaDC and AtSerDC is akin to that of other Group II PLP-dependent amino acid decarboxylases, except for a conspicuous dissimilarity. Unlike other Group II PLP-dependent amino acid decarboxylases, the C-terminal of CsAlaDC and AtSerDC harbors an evident zinc finger structure. We propose that this structure could potentially influence enzyme stability. To test this hypothesis, we truncated the zinc finger structure from both proteins and expressed them. Our findings indicate that, following the excision of the zinc finger structure, CsAlaDC became insoluble, while AtSerDC displayed a similar trait to a certain extent (Supplementary figure 9). Hence, we can establish that the zinc finger structure indeed exerts an influence on the enzyme’s stability. The potential for additional functions necessitates further investigation.

Evolution of AlaDC and SerDC

The structure of CsAlaDC seems like with AtSerDC, however, there are differences in substrate specificity. To clarify the relationship between CsAlaDC and other SerDCs, we analyzed serine decarboxylase-like proteins in Embryophyta that are highly homologous to CsAlaDC and constructed phylogenetic trees to gain insight into their evolutionary relationships (Supplementary figure 3). Based on our experimental findings and evolutionary evidence, we have identified a conserved YPX motif at the substrate binding pocket. For most plants, the residue X in this motif is predominantly Tyr, which corresponds to the serine decarboxylase protein. However, other residues such as Thr, Phe, Val, Ala, and Ile can also occupy this position (Figure 4B and C). Interestingly, serine decarboxylase-like proteins containing YPF motifs, which have the potential to function as alanine decarboxylases, were found to be distributed throughout the phylogenetic tree, including Asteroideae, Ericales, Chenopodiaceae, Poaceae, and Ranunculales (Supplementary figure 3B). However, these proteins were absent in some more recent species. This observation implies that the emergence of alanine decarboxylase in these particular species could be attributed to convergent evolution.

Moreover, we have observed the enrichment of serine decarboxylase-like proteins that possess a YPT motif within the Fabales (Supplementary figure 3B). Prior research has demonstrated that XP_004496485-1, which possesses a YPT motif, displays catalytic efficacy toward the processes of decarboxylation and oxidative deamination of Phe, Met, Leu, and Trp (36). Given that the YPT motif is highly conserved and widely distributed in Fabales, serine decarboxylase-like proteins bearing the YPT motif may have developed a unique substrate specificity in Fabales, beyond their conventional decarboxylation functions, as exemplified by XP_004496485-1 protein mentioned above. We speculate that these proteins may be capable of catalyzing other reactions, potentially involving non-protein amino acids or other substances as substrates.

Applied to improve the synthesis of theanine

Theanine is an important indicator of green tea quality. Therefore, improving the synthesis of theanine in tea plants is the focus of research. In this study, through crystal structure analysis and mutation verification, we found that the catalytic activity of CsAlaDCL110F/P114A is 2.3 times higher than that of the wild-type protein, resulting in a more abundant synthesis of theanine in vitro. This gives us great inspiration to improve the synthesis of EA in tea plants by gene editing, thus increasing the content of theanine in tea plants (Figure 5).

Theanine is highly demanded, by the market, due to its health effects and medicinal value, and as a constituent in food, cosmetics and other fields. To meet the market demand, a variety of methods have been used to acquire theanine, with the main methods including direct extraction, chemical synthesis, biotransformation (microbial fermentation) and plant cell culture. Based on the findings of this study, site-directed mutagenesis can be employed to modify enzymes involved in theanine synthesis. This modification enhances the capacity of bacteria, yeast, model plants, and other organisms to synthesize theanine, thereby facilitating its application in industrial theanine production.

Conclusions

In conclusion, our structural and functional analyses have significantly advanced understanding of the substrate-specific activities of alanine and serine decarboxylases, typified by CsAlaDC and AtSerDC. Critical amino acid residues responsible for substrate selection were identified-Tyr111 in AtSerDC and Phe106 in CsAlaDC-highlighting pivotal roles in enzyme specificity. The engineered CsAlaDC mutant (L110F/P114A) not only displayed enhanced catalytic efficiency but also substantially improved L-theanine yield in a synthetic biosynthesis setup with PsGS or MmGMAS. Our research expanded the repertoire of potential alanine decarboxylases through the discovery of 13 homologous enzyme candidates across embryophytic species and uncovered a special motif present in serine protease-like proteins within Fabale, suggesting a potential divergence in substrate specificity and catalytic functions. These insights lay the groundwork for the development of industrial biocatalytic processes, promising to elevate the production of L-theanine and supporting innovation within the tea industry.

Materials and Methods

Plants materials

Tobacco (Nicotiana benthamiana) plants were grown in a controlled chamber, under a 16-h light and 8-h dark photoperiod at 25 ℃. Leaves of 5-week-old tobacco plants were used for transient transformation, mediated by Agrobacterium tumefaciens strain GV3101.

Gene cloning and protein expression

The coding sequences of AtSerDC and CsAlaDC, amino acid residues 66-482 and 61-478, respectively, were amplified using PCR and then ligated into the Nde Ι and Xho I restriction sites of the pET-28a and pET-22b vector, respectively, providing the recombinant vector pET-28a-AtSerDC and pET-22b-CsAlaDC. The cDNAs encoding CsSerDC were amplified using PCR and then ligated into the Nde Ι and Xho I restriction sites of the pET-28a vector. Gene-specific primers were listed in Supplementary table 2. The recombinant plasmid was transformed into E. coli BL21 (DE3) competent cells. Positive transformants were grown in a 5 mL LB medium containing 30 μg/mL kanamycin or 50 μg/mL ampicillin at 37 °C overnight and then subcultured into an 800 mL LB medium containing the corresponding antibiotic. Protein expression was induced by the addition of 0.2 mM isopropyl-β-d-thiogalactoside (IPTG) for 20 h at 16 ℃ when the optical density (OD) at 600 nm reached 0.6-0.8, harvested by centrifugation at 4 °C and 4,000 rpm for 30 min.

Protein production and crystallization

The cell pellet was suspended in 30 mL of lysis buffer (20 mM HEPES, pH 7.5, 200 mM NaCl, 0.1 mM PLP), disrupted by High-Pressure Homogenizer, and then centrifuged at 16,000 rpm for 30 min at 4 °C to remove the cell debris. The supernatants were purified with a Ni-Agarose resin column followed by size-exclusion chromatography. Before crystallization, purified proteins were concentrated at 10 mg/mL. Crystallization conditions were screened by the sitting-drop vapor diffusion method using the reservoir solutions supplied in commercially available screening kits (Crystal Screen, Crystal Screen 2, PEGRx 1, 2, and SaltRx 1, 2). A droplet made by mixing 1.0 μL of purified AtSerDC or CsAlaDC (10 mg/mL) with an equal volume of a reservoir solution was equilibrated against 100 μL of the reservoir solution at 16℃. The crystal of AtSerDC was obtained using buffer pH 8.0 containing 20% (w/v) PEG400 as a precipitate and 0.2 M CaCl2. The crystal of the CsAlaDC-EA complex was obtained at pH 7.5 containing 2.6 M sodium acetate. The crystal of CsAlaDC was obtained at pH 6.0 containing 3.5 M sodium formate.

Data collection and processing

Crystals were grown by sitting-drop vapor diffusion method at 16 °C. The volume of the reservoir solution was 100 µL and the drop volume was 2 µL, containing 1 µL of protein sample and 1 µL of reservoir solution. The reservoir solution of AtSerDC contained 0.1 M Tris-HCl (pH 8.0), 0.2 M CaCl2, and 20% PEG 400. The reservoir solution of CsAlaDC contained 0.1 M HEPES (pH 7.5), 2.6 M sodium acetate or 0.1 M Bis-Tris (pH 6.0), and 3.5 M sodium formate. Crystals grew in 3-5 days using a protein concentration of 10 mg/mL. Diffraction data were collected at Shanghai Synchrotron Radiation Facility (China). The collected data sets were indexed, integrated, and scaled using the HKL3000 software package. The structure of AtSerDC was solved by molecular replacement using the structure of HisDC (PDB code: 7ERV [Photobacterium phosphoreum]) as the model, and utilizing AtSerDC as a molecular displacement model for both CsAlaDC and the CsAlaDC-EA complex. The AtSerDC, CsAlaDC, and CsAlaDC-EA complex exhibit resolutions of 2.85 Å, 2.50 Å, and 2.60 Å, respectively. The statistics for data collection and processing are summarized in Supplementary table 1.

Enzyme activity assays

Decarboxylase activity was measured by detecting products (EA or ethanolamine) in Waters Acquit ultraperformance liquid chromatography (UPLC) system (16, 37). The 100 μL reaction mixture, containing 20 mM substrate (Ala or Ser), 100 mM potassium phosphate, 0.1 mM PLP, and 0.025 mM purified enzyme, was prepared and incubated at standard conditions (45 °C and pH 8.0 for CsAlaDC, 40 °C and pH 8.0 for AtSerDC for 10 min). Then, the reaction was stopped with 20 μL of 10% trichloroacetic acid. The product was derivatized with 6-aminoquinolyl-N-hydroxy-succinimidyl carbamate (AQC) and subjected to analysis by UPLC. All enzymatic assays were performed in triplicate.

The detection methodology for theanine production remains consistent with the aforementioned approach. The 100 μL reaction mixture, containing 20 mM Ala, 45 mM Glu, 50 mM HEPES (pH 7.5), 0.1 mM PLP, 30 mM MgCl2, 10 mM ATP, 0.03 mM PsGS/MmGMAS, and 0.025 mM CsAlaDC/CsAlaDC L110F/CsAlaDC L110F/P114A, was prepared and incubated at standard conditions for 1 h. Subsequently, the reaction was terminated via immersion of the reaction vessel in a metal bath at 96 °C for 3 min (31). Theanine was derivatized with AQC and subjected to analysis by UPLC. All enzymatic assays were performed in triplicate.

Site-directed mutagenesis

Site-directed mutagenesis experiment was conducted using a PCR method from the wild-type construct pET-28a-AtSerDC and pET-22b-CsAlaDC, respectively. Dpn I endonucleases were used to digest the parental DNA template. The reaction mixture was used to transform E. coli DH5α competent cells and the plasmids from positive strains were extracted to E. coli BL21 (DE3) for protein expression, purification, and analysis of the enzymatic activity.

In vivo enzyme activity assay in N. benthamiana

The amplified PCR products were fused to the plant expression vector, pCAMBIA1305. Linearization was conducted by restriction digest with Spe I and BamH I. The recombinant colonies were selected for PCR validation on the appropriate antibiotics plate. After validation, the plasmids were electroporated into Agrobacterium tumefaciens strain GV3101(pSoup-p19). Empty pCAMBIA1305 vector, with an intron containing the GFP gene, was treated as the control.

Agrobacterium transient expression assays were performed on 5-week-old N. benthamiana plants. Agrobacterium tumefaciens strain GV3101 (pSoup-p19), carrying the above-described vectors, were cultured in Luria Bertani (LB) medium, containing appropriate antibiotics, at 28 °C. When the absorbance of bacteria colonies reached OD600 = 0.6-0.8, bacterial cells were collected and resuspended in MMA solution (10 mM MgCl2, 10 mM 2-(N-morpholino) ethane sulfonic acid (MES), pH 5.6). After the OD600 of the resuspended bacterial solution was adjusted to approx. 1.0, acetosyringone (AS) was added with a final concentration of 200 µM, and this solution was then incubated, at room temperature, for at least 3 h in darkness. Next, cell suspensions were infiltrated into N. benthamiana leaves with a needle-free syringe. These N. benthamiana leaves were then collected 3 days post infiltration, frozen in liquid nitrogen and stored at –80 °C. Internal EA in N. benthamiana leaves was extracted as previously described (18), and then the solvent was subjected to gas chromatography-mass spectrometry GS-MS system.

Transcript level analysis in N. benthamiana

Total RNA was isolated from samples using the RNAprep Pure Plant Kit (Tiangen, Beijing, China), according to the manufacturer’s protocol. The cDNAs were synthesized using TransScript One-Step gDNA Removal and cDNA Synthesis SuperMix Kit (TransGen Biotech, Beijing, China). The qRT-PCR assays were performed, as previously described (38). Primers used for qRT-PCR assays were listed in Supplementary table 3 and the qRT-PCR was run on a Bio-Rad CFX96TM RT PCR detection system and CFX Manager Software. Each reaction reagent (20 μL) contained 0.4 μL forward and reverse primers (10 μM), 2 μL cDNA (200±5 ng/μL), 10 μL SYBR Green Supermix (Vazyme, Nanjing, China) and 6.2 μL double-distilled water. Reaction was performed by a two-step method: 95 ℃ for 5 min; 40 cycles of 95 ℃ for 10 s; and 60 ℃ for 30 s. The glyceraldehyde-3-phosphate dehydrogenase (GAPDH) gene was used for internal normalization in each RT-PCR, and the 2−ΔΔCT method (39) was used to calculate the relative gene expression. All samples were performed with three replicates.

Multiple sequence alignment

MUSCLE was used to generate the protein multiple sequence alignment of amino acid decarboxylases with default settings (40). ESPript 3.x was used to display the multiple sequence alignment (41).

Phylogenetic analysis

The AtSerDC was used as a query for searching its homologs in Embryophyta using BLASTP (42) (version 2.13.0). To fetch homologs of AtSerDC in distant species, we performed a two-round blast. We first search for the best hit in the given order under Embryophyta. Then we blast each species under a provided order using the order best hit as a query and as many as 10 hits were kept. The sequences with e values <= 0.001 were removed. We further dropped sequences with lengths less than 200 or larger than 800 amino acids. Then multiple sequence alignment was performed via clustalo (version 1.2.4) with default settings on these filtered sequences (43). The NJ algorithm implemented in clustalw (version 2.1) was adopted to build the phylogenetic tree (44). The R package ggtree (version 3.2.1) and universalmotif (version 1.12.4) were used for tree and motif visualization (45).

Author contributions

W.G, Z.Z and X.W designed research; H.W, B.Z, S.Q and C.D performed research; H.W and B.Z analyzed data; H.W, B.Z, W.G, and Z.Z wrote the paper.

Competing interest statement

The authors declare no competing financial interests.

Classification

Biochemistry (major), Plant Sciences (minor)